---
title: "Other benchmarks"
format: html
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE
)
```

```{r setup, eval=TRUE, echo=FALSE}
suppressMessages(
  {
    library(broadcast)
    library(bench)
    library(tinyplot)
    library(ggplot2)
    library(abind)
  }
)

```


&nbsp;

# Introduction

This page benchmarks some of the functions from 'broadcast' with some near-equivalent functions from other packages. The code is given here also.

&nbsp;

# `abind::abind()`

In this section, te performance of the [bind_array()](/man/bind_array.qmd) function from 'broadcast' is compared to the performance of the `abind()` function from the 'abind' package.

&nbsp;

The following code was used:


```{r echo = TRUE, eval=FALSE}
n <- 110L
nms <- function(n) sample(letters, n, TRUE)
x <- array(as.double(1:25), c(n, n, n))
y <- array(as.double(-1:-25), c(n, n, n))
dimnames(x) <- lapply(dim(x), nms)
dimnames(y) <- lapply(dim(y), nms)
input <- list(x, y, x)

gc()
bm_abind <- bench::mark(
  abind = abind::abind(input, along = 2),
  broadcast = bind_array(input, 2),
  min_iterations = 100,
  check = FALSE # because abind adds empty dimnames
)
summary(bm_abind)
plot(bm_abind)

```

And here are the results:

```{r fig.width=8, echo=FALSE}
load("benchmarks/bm_abind.RData")
summary(bm_abind)
plot(bm_abind)
```


Clearly, the [bind_array()](/man/bind_array.qmd) function from 'broadcast' is about 2 to 3 times faster than the `abind()` function from the 'abind' package. It is also about 2 times more memory efficient.

&nbsp;

# `Rfast::Outer()`

An outer computation is a special case of broadcasting, namely a broadcasting computation between a row-vector and a column-vector. The `outer()` function from base 'R' is too slow and consumes too much memory to provide any meaningful benchmark. But the 'Rfast' package provides a very fast implementation of the `outer()` function. It may be interesting how broadcasted operations hold up to the famously fast 'Rfast' package.

Here the outer-sum between a row-vector `x` and column-vector `y` (both have 9000 elements) is computed using `Rfast::outer()` and `broadcast::bc.num()`, and their speeds and memory consumption are compared.

The following code was used:

```{r echo = TRUE, eval=FALSE}

n <- 9e3
x <- array(rnorm(10), c(1, n))
y <- array(rnorm(10), c(n, 1))
gc()
bm_outer <- bench::mark(
  Rfast = Rfast::Outer(x, y, "+"),
  broadcast = bc.num(x, y, "+"),
  min_iterations = 100
)
summary(bm_outer)
plot(bm_outer)

```

And here are the results:

```{r fig.width=8, echo=FALSE}
load("benchmarks/bm_outer.RData")
summary(bm_outer)
plot(bm_outer)
```


It seems that the implementations of 'broadcast' and the blazingly fast 'Rfast' package reach similar speeds and use the same amount of memory.

Note, however, that `Rfast::Outer()` unfortunately only supports numeric vectors, and does not provide higher-dimensional broadcasting. 'broadcast', on the other hand, supports all atomic types as well as the `list` recursive type, and supports arrays of any dimensions up to 16 dimensions.


&nbsp;

# `%r+%` operator from 'collapse'


The impressive 'collapse' package supports a large set of blazingly fast functions for a large variety of tasks. One of these is the `x %r% v` operator. Given a matrix `x` and a vector `v`, `x %r+% v` will add `v` to every row of `x`. Using this function in this way is equivalent to the [bc.num()](/man/bc.d.qmd) function, using a column-vector for `v`.

Here these 2 approaches are benchmarked.

The code used was as follows:

```{r echo=TRUE, eval=FALSE}

n <- 8e3
x <- matrix(rnorm(10), n, n)
v <- array(rnorm(10), c(1, n))
bm_collapse_row <- bench::mark(
  collapse = x %r+% v,
  broadcast = bc.num(x, v, "+"),
  min_iterations = 100
)
summary(bm_collapse_row)
plot(bm_collapse_row)
```



And here are the results:

```{r fig.width=8, echo=FALSE}
load("benchmarks/bm_collapse_row.RData")
summary(bm_collapse_row)
plot(bm_collapse_row)
```

The 'collapse' package is slightly faster than 'broadcast' in this case. This does show how super fast 'collapse' tru;ly is.


