---
title: "Other benchmarks"
format: html
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE
)
```

```{r setup, eval=TRUE, echo=FALSE}
suppressMessages(
  {
    library(broadcast)
    library(bench)
    library(tinyplot)
    library(ggplot2)
    library(abind)
    library(fontawesome)
  }
)

```


&nbsp;

# Introduction

This page benchmarks some of the functions from 'broadcast' with some near-equivalent functions from other packages. The code is given here also.

The 'benchmark' package was used for measuring speed and memory usage, and for producing the figures showing the results.

The benchmarks were all run on the same computer (processor: 12th Gen Intel(R) Core(TM) i5-12500H @ 2.50 GHz) with 32GB of RAM and running the Windows 11 OS (64 bit).

`r fa("r-project")` version 4.4.0 with 'Rstudio' version 2024.12.1 was used to run the `r fa("r-project")` code.

The various comparisons are split over several sections. The code used to run the benchmarks is given in each section, just before the results.

&nbsp;

# `abind::abind()`

In this section, te performance of the [bind_array()](/man/bind_array.qmd) function from 'broadcast' is compared to the performance of the `abind()` function from the 'abind' package.

&nbsp;

The following code was used:


```{r echo = TRUE, eval=FALSE}
n <- 110L
nms <- function(n) sample(letters, n, TRUE)
x <- array(as.double(1:25), c(n, n, n))
y <- array(as.double(-1:-25), c(n, n, n))
dimnames(x) <- lapply(dim(x), nms)
dimnames(y) <- lapply(dim(y), nms)
input <- list(x, y, x)

gc()
bm_abind <- bench::mark(
  abind = abind::abind(input, along = 2),
  broadcast = bind_array(input, 2),
  min_iterations = 100,
  check = FALSE # because abind adds empty dimnames
)
summary(bm_abind)
plot(bm_abind)

```

And here are the results:

```{r fig.width=8, echo=FALSE}
load("benchmarks/bm_abind.RData")
summary(bm_abind)
plot(bm_abind)
```


Clearly, the [bind_array()](/man/bind_array.qmd) function from 'broadcast' is about 2 to 3 times faster than the `abind()` function from the 'abind' package. It is also about 2 times more memory efficient.

&nbsp;

# `Rfast::Outer()`

An outer computation is a special case of broadcasting, namely a broadcasting computation between a row-vector and a column-vector. The `outer()` function from base 'R' is too slow and consumes too much memory to provide any meaningful benchmark. But the 'Rfast' package provides a very fast implementation of the `outer()` function. It may be interesting how broadcasted operations hold up to the famously fast 'Rfast' package.

Here the outer-sum between a row-vector `x` and column-vector `y` (both have 9000 elements) is computed using `Rfast::outer()` and `broadcast::bc.d()`, and their speeds and memory consumption are compared.

The following code was used:

```{r echo = TRUE, eval=FALSE}

n <- 9e3
x <- array(rnorm(10), c(1, n))
y <- array(rnorm(10), c(n, 1))

gc()
bm_outer <- bench::mark(
  Rfast = Rfast::Outer(x, y, "+"),
  broadcast = bc.d(x, y, "+"),
  min_iterations = 100
)
summary(bm_outer)
plot(bm_outer)

```

And here are the results:

```{r fig.width=8, echo=FALSE}
load("benchmarks/bm_outer.RData")
summary(bm_outer)
plot(bm_outer)
```


It seems that the implementations of 'broadcast' and the blazingly fast 'Rfast' package reach similar speeds and use the same amount of memory.

Note, however, that `Rfast::Outer()` unfortunately only supports numeric vectors, and does not provide higher-dimensional broadcasting. 'broadcast', on the other hand, supports all atomic types as well as the `list` recursive type, and supports arrays of any dimensions up to 16 dimensions.


&nbsp;

# `%r+%` operator from 'collapse'


The impressive 'collapse' package supports a large set of blazingly fast functions for a large variety of tasks. One of these is the `x %r% v` operator. Given a matrix `x` and a vector `v`, `x %r+% v` will add `v` to every row of `x`. Using this function in this way is equivalent to the [bc.d()](/man/bc.d.qmd) function, using a row-vector for `v`.

Here these 2 approaches are benchmarked.

The code used was as follows:

```{r echo=TRUE, eval=FALSE}

n <- 8e3
x <- matrix(rnorm(10), n, n)
v <- array(rnorm(10), c(1, n))

gc()
bm_collapse_row <- bench::mark(
  collapse = x %r+% v,
  broadcast = bc.d(x, v, "+"),
  min_iterations = 100
)
summary(bm_collapse_row)
plot(bm_collapse_row)
```



And here are the results:

```{r fig.width=8, echo=FALSE}
load("benchmarks/bm_collapse_row.RData")
summary(bm_collapse_row)
plot(bm_collapse_row)
```

The 'collapse' package is slightly faster than 'broadcast' in this case. This does show how super fast 'collapse' truly is.

&nbsp;


# Base 'R' replication

Here replicating array dimensions using base 'R' is benchmarked against broadcasting.

The following code was used:

```{r echo=TRUE, eval=FALSE}

n <- 450
x <- array(rnorm(10), c(1, n, 1))
y <- array(rnorm(10), c(n, 1, n))

gc()
bm_base <- bench::mark(
  base = x[rep(1, n), , rep(1, n)] + y[, rep(1, n), ],
  broadcast = bc.d(x, y, "+"),
  min_iterations = 100
)
summary(bm_base)
plot(bm_base)

```


And here are the results:

```{r fig.width=8, echo=FALSE}
load("benchmarks/bm_base.RData")
summary(bm_base)
plot(bm_base)
```


'broadcasting' is 5 to 5.5 times (!) faster than replicating array dimensions, and uses approximately 2 times less memory.

