---
title: "Other benchmarks"
format: html
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE
)
```

```{r setup, eval=TRUE, echo=FALSE}
suppressMessages(
  {
    library(broadcast)
    library(bench)
    library(tinyplot)
    library(ggplot2)
    library(abind)
    library(fontawesome)
  }
)

```


&nbsp;

# Introduction

This page benchmarks some of the functions from the latest version (`0.1.6.1`) of 'broadcast' with some near-equivalent functions from base `r fa("r-project")` and other packages. The code is given here also.

Because all these benchmarks are within `r fa("r-project")`, not only speed but also memory usage is compared.

&nbsp;

## Why care about speed and memory usage?

Faster code means less waiting for the user, which is less annoying. Moreover, faster code is (generally speaking) more efficient code - i.e. you don't have to run your computer overnight (or at least less hours) - which reduces electricity usage. Using less electricity is better for your wallet and better for the environment.

Code that uses less memory when run means less resources (i.e. less memory) is required. And using less memory reduces the risk of running out of usable memory; when you run out of usable memory, your code may stop running with the following error:

```{r eval=FALSE, echo=TRUE}
Error: cannot allocate vector of size
```

You may want to prevent that. And without more memory-efficient code you'll be forced to buy more memory (which costs money).


&nbsp;

## Materials used

The 'benchmark' package was used for measuring speed and memory usage, and for producing the figures showing the results.

The benchmarks were all run on the same computer (processor: 12th Gen Intel(R) Core(TM) i5-12500H @ 2.50 GHz) with 32GB of RAM and running the Windows 11 OS (64 bit).

`r fa("r-project")` version 4.4.0 with 'Rstudio' version 2024.12.1 was used to run the `r fa("r-project")` code.

The various comparisons are split over several sections. The code used to run the benchmarks is given in each section, just before the results.


&nbsp;


# Base `r fa("r-project")` replication

Here replicating array dimensions using base `r fa("r-project")` is benchmarked against broadcasting.

The following code was used:

```{r echo=TRUE, eval=FALSE}

n <- 450
x <- array(rnorm(10), c(1, n, 1))
y <- array(rnorm(10), c(n, 1, n))

gc()
bm_base <- bench::mark(
  base = x[rep(1, n), , rep(1, n)] + y[, rep(1, n), ],
  broadcast = bc.d(x, y, "+"),
  min_iterations = 100
)
summary(bm_base)
plot(bm_base)

```


And here are the results:

```{r fig.width=8, echo=FALSE}
load("benchmarks/bm_base.RData")
summary(bm_base)
plot(bm_base)
```


'broadcasting' is **5 to 5.5** times (!) faster than replicating array dimensions, and uses approximately 2 times less memory.

&nbsp;

# `sweep()`

Here the `sweep()` function from base `r fa("r-project")` is benchmarked against broadcasting.

The following code was used:

```{r echo=TRUE, eval=FALSE}

n <- 2000
x <- matrix(rnorm(n), n, n)
cm <- array(colMeans(x), c(1, n))

gc()
bm_sweep <- bench::mark(
  sweep = sweep(x, 2, cm, "=="),
  broadcast = bc.d(x, cm, "=="),
  min_iterations = 100
)
summary(bm_sweep)
plot(bm_sweep)
```


And here are the results:

```{r fig.width=8, echo=FALSE}
load("benchmarks/bm_sweep.RData")
summary(bm_sweep)
plot(bm_sweep)
```


'broadcasting' is about **3** (!) times faster than using `sweep()`, and uses approximately **5** (!) times less memory.

&nbsp;

# `abind::abind()` - binding multi-dimensional arrays

In this section, the performance of the [bind_array()](/man/bind_array.qmd) function from 'broadcast' is compared to the performance of the `abind()` function from the 'abind' package, for binding 3 moderately sized multi-dimensional arrays.

&nbsp;

The following code was used:


```{r echo = TRUE, eval=FALSE}
n <- 110L
nms <- function(n) sample(letters, n, TRUE)
x <- array(as.double(1:25), c(n, n, n))
y <- array(as.double(-1:-25), c(n, n, n))
dimnames(x) <- lapply(dim(x), nms)
dimnames(y) <- lapply(dim(y), nms)
input <- list(x, y, x)

gc()
bm_abind <- bench::mark(
  abind = abind::abind(input, along = 2),
  broadcast = bind_array(input, 2),
  min_iterations = 100,
  check = FALSE # because abind adds empty dimnames
)
summary(bm_abind)
plot(bm_abind)

```

And here are the results:

```{r fig.width=8, echo=FALSE}
load("benchmarks/bm_abind_arrays.RData")
summary(bm_abind)
plot(bm_abind)
```


Clearly, the [bind_array()](/man/bind_array.qmd) function from 'broadcast' is about about 5 times faster than the `abind()` function from the 'abind' package. It is also about 4 times more memory efficient.

&nbsp;


# `abind::abind()` - binding vectors

In this section, the performance of the [bind_array()](/man/bind_array.qmd) function from 'broadcast' is compared to the performance of the `abind()` function from the 'abind' package, for binding many (100) 1d arrays.

&nbsp;

The following code was used:


```{r echo = TRUE, eval=FALSE}
input <- rep(list(array(1:1000, c(1, 1000))), 100)
gc()
bm_abind <- bench::mark(
  abind = abind::abind(input, along = 2),
  broadcast = bind_array(input, 2),
  min_iterations = 100,
  check = FALSE # because abind adds empty dimnames
)
summary(bm_abind)
ggplot2::autoplot(bm_abind)

```

And here are the results:

```{r fig.width=8, echo=FALSE}
load("benchmarks/bm_abind_vectors.RData")
summary(bm_abind)
plot(bm_abind)
```


Clearly, the [bind_array()](/man/bind_array.qmd) function from 'broadcast' is about about **25** (!!!) times faster than the `abind()` function from the 'abind' package. It is also about **100** (!!!) times more memory efficient.

&nbsp;

# `Rfast::Outer()`

An outer computation is a special case of broadcasting, namely a broadcasting computation between a row-vector and a column-vector.

Base `r fa("r-project")` has the `outer()` function to perform outer computations. But the `outer()` function from base `r fa("r-project")` uses **soo much memory**, that no meaningful benchmark can be made against the methods provided  by the 'broadcast' package.

The 'Rfast' package, however, provides a very fast and memory-efficient implementation of the `outer()` function. It may be interesting how broadcasted operations hold up to the famously fast 'Rfast' package.

Here the outer-sum between a row-vector `x` and column-vector `y` (both have 9000 elements) is computed using `Rfast::outer()` and `broadcast::bc.d()`, and their speeds and memory consumption are compared.

The following code was used:

```{r echo = TRUE, eval=FALSE}

n <- 9e3
x <- array(rnorm(10), c(1, n))
y <- array(rnorm(10), c(n, 1))

gc()
bm_outer <- bench::mark(
  Rfast = Rfast::Outer(x, y, "+"),
  broadcast = bc.d(x, y, "+"),
  min_iterations = 100
)
summary(bm_outer)
plot(bm_outer)

```

And here are the results:

```{r fig.width=8, echo=FALSE}
load("benchmarks/bm_outer.RData")
summary(bm_outer)
plot(bm_outer)
```


It seems that the implementations of 'broadcast' and the blazingly fast 'Rfast' package reach similar speeds and use the same amount of memory.

Note, however, that `Rfast::Outer()` unfortunately only supports numeric vectors, and does not provide higher-dimensional broadcasting. 'broadcast', on the other hand, supports all atomic types as well as the `list` recursive type, and supports arrays of any dimensions up to 16 dimensions.


&nbsp;

# `%r+%` operator from 'collapse'


The impressive 'collapse' package supports a large set of blazingly fast functions for a large variety of tasks. One of these is the `x %r% v` operator. Given a matrix `x` and a vector `v`, `x %r+% v` will add `v` to every row of `x`. Using this function in this way is equivalent to the [bc.d()](/man/bc.d.qmd) function, using a row-vector for `v`.

Here these 2 approaches are benchmarked.

The code used was as follows:

```{r echo=TRUE, eval=FALSE}

n <- 8e3
x <- matrix(rnorm(10), n, n)
v <- array(rnorm(10), c(1, n))

gc()
bm_collapse_row <- bench::mark(
  collapse = x %r+% v,
  broadcast = bc.d(x, v, "+"),
  min_iterations = 100
)
summary(bm_collapse_row)
plot(bm_collapse_row)
```



And here are the results:

```{r fig.width=8, echo=FALSE}
load("benchmarks/bm_collapse_row.RData")
summary(bm_collapse_row)
plot(bm_collapse_row)
```

The 'collapse' package is slightly faster than 'broadcast' in this case. This does show how super fast 'collapse' truly is.

&nbsp;

# Sub-setting with nested list vs dimensional list

Performing tasks with dimensional lists is often faster and more convenient with dimensional list than with nested lists.

One example is sub-setting.  
In the benchmarks below, 2 equivalent sub-set operations are compared:

 - selecting `x[[i]][[1L]]` for `i = 1:length(x)` in a nested list (this requires a for-loop).
 - selecting `x2[, 1L]` in a dimensional list (this is a vectorized operation).

where `x2 = cast_hier2dim(x, in2out = FALSE)`.

The following code is used:

```{r echo=TRUE, eval=FALSE}

x <- lapply(1:200, function(x) sample(1:200))
x <- rep(list(x), 200)
x2 <- cast_hier2dim(x, in2out = FALSE)
bm_hier_vs_dim <- bench::mark(
  nested = for(i in 1:length(x)) x[[i]][[1L]],
  dimensional = x2[, 1L , drop = FALSE],
  min_iterations = 200,
  check = FALSE
)
summary(bm_hier_vs_dim)
plot(bm_hier_vs_dim)

```


```{r fig.width=8, echo=FALSE}
load("benchmarks/bm_hier_vs_dim.RData")
summary(bm_hier_vs_dim)
plot(bm_hier_vs_dim)
```

As shown in the benchmarks, sub-setting with a dimensional list is **hundreds of times** faster than doing the equivalent sub-setting operation in a nested list.

&nbsp;

&nbsp;
