---
title: "Table of Contents"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo=FALSE, eval=TRUE}
library(broadcast) |> suppressMessages()
library(fontawesome) |> suppressWarnings()
```



The following Guides and Vignettes are available:

 - [Quick-Start Guide](b_quickstart.qmd): The Quick-Start Guide is the first thing to read for new users, and gives a brief explanation on how to use the 'broadcast' package, without getting lost in the details.
 - [Prerequisites and Installation](installation.qmd): This guide explains the requirements to use the 'broadcast' package, and how to install it.
 - [Functions Overview](functions.qmd): This guide simply lists the functions available in the broadcast package.
 - [Broadcasting Explained](broadcasting_explained.qmd): This Guide gives a very detailed explanation of how exactly broadcasting works in the "broadcast" package. Read this *after* you've read the Quick-Start Guide.
 - The [Operator Overloading](operators.qmd), [Binding Broadcasted](binding.qmd), and [List Casting](casting.qmd) Vignettes explain the overloaded operators & the `bc.*` functions, broadcasted array binding, and list casting methods, respectively.
 - [Practical Applications](practical_applications.qmd): This Guide gives some practical applications wherein the broadcasted methods might be used.
 - [Technical Details](technical_details.qmd): This Guide gives some technical details; not relevant for the average use.

