---
title: "Applications in Data Wrangling"
format:
  html:
    number-sections: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo=FALSE, eval=TRUE}
library(broadcast) |> suppressMessages()
library(magrittr) |> suppressMessages()
library(fontawesome) |> suppressWarnings()
set.seed(1L)
```


&nbsp;


Broadcasting comes up frequent enough in real world  data wrangling problems. This page gives a few examples of these.

&nbsp;

# Sweep

Anytime you would normally use `sweep()`, you can now use broadcasting, which is faster and more memory-efficient.

Consider for example the following matrix:

```{r}
mat <- matrix(rnorm(20), nrow = 5)
bcr(mat) <- TRUE # `bcr` is short-hand for `broadcaster`
print(mat)
```

We want to scale each column of this matrix, by subtracting its mean and dividing by its standard deviation.

This can be done using `sweep()` like so:

```{r}
scaled <- sweep(mat, 2, colMeans(mat), FUN = "-")
scaled <- sweep(scaled, 2, matrixStats::colSds(mat), FUN = "/")
print(scaled)
```

But it can be done much faster and more memory-efficiently with broadcasting like so:

```{r}
means <- matrix(colMeans(mat), nrow = 1L)
sds <- matrix(matrixStats::colSds(mat), nrow = 1L)
bcr(means) <- bcr(sds) <- TRUE
scaled <- (mat - means) / sds
print(scaled)
```

The larger the matrix `mat` becomes, the more advantageous it becomes to use broadcasting rather than sweeping.

&nbsp;

# Bind arrays

The `abind()` function, from the package of the same name, allows one to bind arrays along any arbitrary dimensions (not just along rows or columns).  
Unfortunately, `abind()` does not support broadcasting, which can lead to frustrations such as the following:

```{r echo=TRUE, eval=FALSE}
(x <- array(1:27, c(3,3,3)))
(y <- array(1L, c(3,3,1)))
abind::abind(x, y, along = 2)

#> Error in abind(x, y, along = 2) : 
#>  arg 'X2' has dims=3, 3, 1; but need dims=3, X, 3
```

Here, `abind()` is complaining about the dimensions not fitting perfectly.  
But intuitively, binding `x` and `y` should be possible, with dimension `3` from array `y` being broadcasted to size 3.

The [bind_array()](/man/bind_array.qmd) function provided by the 'broadcast' package can bind the arrays without problems:

```{r echo=TRUE, eval=TRUE}
x <- array(1:27, c(3,3,3))
y <- array(1L, c(3,3,1))
bind_array(list(x, y), 2)
```

[bind_array()](/man/bind_array.qmd) is also considerably faster and more memory efficient than `abind()`.

&nbsp;


# Compute on all possible pairs

Suppose you have 2 vectors of character strings, and you want to calculate the distance (i.e. dissimilarity) all possible pairs of these strings.

In base `r fa("r-project")`, this would require a either a loop (which is slow), or repeating the vectors several times (which requires more memory), or use of the `outer()` function (which is both slow **and** requires lots of memory).

The 'broadcasted way to do this, is to make the vectors orthogonal, and then calculate the distances between the strings of the orthogonal vectors - this is both fast and memory efficient.

For example:

```{r}
x <- array(month.name, c(12, 1))
y <- array(month.abb, c(1, 12))
out <- bc.str(x, y, "levenshtein")
dimnames(out) <- list(month.name, month.abb)
print(out)

```

&nbsp;

# Modify Array Along a Dimension

## Along columns

Consider the following array:

```{r}
x <- array(sample(0:9), c(3,3,3))
print(x)
```

Suppose one wishes to add `10` to all values of the second column, and `100` to all values of the third column.

This can be done with 'broadcast', as follows:  
First, create a modification broadcaster vector, whose dimensional direction is aligned with - in this case - the columns; this can be done easily using, for example, `vector2array(v, direction = direction, ndim  = ndim, broadcaster = TRUE)` (see the help page of [vector2array()](/man/vector2array.qmd) for details).  
Second, make `x` a broadcaster, and perform the broadcasted addition between the array and directional vector.

We don't want to change the first column, so we add `0` to the first column.  

So we can do this as follows:

```{r}
v <- vector2array(c(0, 10, 100), 2, 2, TRUE)
print(v)
bcr(x) <- TRUE

x + v

```

&nbsp;

## Along rows

Suppose that, instead of adding `{0, 10, 100}` to the 3 columns, we wish to add these to the rows instead.

This again can by creating a modifcation broadcaster vector - this time aligned with the rows.  
Note that in the 'broadcast' package, a vector without dimensions is interpreted as a vector aligned with rows (confusingly called a "column vector" in mathematics).
Like so:

```{r}
v <- c(0, 10, 100) # effectively the same as array(c(0, 10, 100), c(3, 1))
print(v)
bcr(x) <- bcr(v) <- TRUE

x + v

```

&nbsp;

## Along layers

As a final example, we want to add `{0, 10, 100}` to the layers (3rd dimension) of `x`.  
As you may have guessed, one can do so by creating a modification broadcaster vector aligned with the layers, and adding that vector to the array, like so:

```{r}
v <- vector2array(c(0, 10, 100), 3, 3, TRUE)
print(v)
bcr(x) <- TRUE

x + v

```


&nbsp;

# Grouped Broadcasting

## Casting with equal group sizes

We have a matrix of points (ranging from 0 to 100) students (n = 3) have achieved in for 2 homework exercises:

```{r}
x <- cbind(
  student = rep(1:3, each = 2),
  homework = rep(1:2, 3),
  points = sample(0:100, 6)
)
print(x)
```

However, the teacher has realised that the second homework assignment was way more difficult than the first, and thus decided the second home work assignments should be weigh more - 2 times more to be precise.

Thus the points for homework assignment 2 should be multiplied by 2. There are various ways to do this. For the sake of demonstration, an approach using [acast()](/man/acast.qmd) is shown here, as this is an example of a **grouped** broadcasted operation. 

'broadcast' allows users to cast subsets of an array onto a new dimension, based on some grouping factor - in this case the homework ID is the grouping factor, and the following will do the job:

```{r}
margin <- 1L # we cast from the rows, so margin = 1
grp <- as.factor(x[, "homework"]) # factor to define which rows belongs to which group
levels(grp) <- c("assignment 1", "assignment 2") # names for the new dimension
out <- acast(x, margin, grp) # casting is performed here
bcr(out) <- TRUE
print(out)
```

Notice that the dimension-names of the new dimension (dimension 3) are equal to `levels(grp)`.

With the group-cast array, one can use broadcasting to easily do things like multiply the values in each group with a different value.  
In this case, we need to multiply the values for assignment 2 with 2, and leave the rest as-is.

Like so:

```{r}

# create the multiplication factor array:
mult <- array(
  1,
  dim = c(1, 3, 2),
  dimnames = list(
    NULL,
    c("mult_id", "mult_homework", "mult_points"),
    c("assignment 1", "assignment 2")
  )
)
mult[, "mult_points", c("assignment 1", "assignment 2")] <- c(1, 2)
bcr(mult) <- TRUE
print(mult)

# grouped broadcasted operation:
out2 <- out * mult
dimnames(out2) <- dimnames(out)
print(out2)

```

Now the array needs to be reverse-cast back to its original shape.  
Reverse-casting an array can be done be combining `asplit()` with [bind_array()](/man/bind_array.qmd):

```{r}
asplit(out2, ndim(out2)) |> bind_array(along = margin, name_along = FALSE)
```

...though the order of, in this case, the rows (because `margin = 1`) will not necessarily be the same as the original array.


&nbsp;


## Casting with unequal group sizes

The casting arrays also works when the groups have unequal sizes, though there are a few things to keep in mind.

Let's start with a different input array:

```{r}
x <- cbind(
  id = c(rep(1:3, each = 2), 1),
  grp = c(rep(1:2, 3), 2),
  val = rnorm(7)
)
print(x)
```

Once again, the [acast()](/man/acast.qmd) to fill the gaps, otherwise an error is called.

Thus one can cast in this case like so:

```{r}
grp <- as.factor(x[, 2])
levels(grp) <- c("a", "b")
margin <- 1L
out <- acast(x, margin, grp, fill = TRUE)
print(out)
```

Notice that some values are missing ( `NA` ); if some groups have unequal number of elements, [acast()](/man/acast.qmd) needs to fill the gaps with missing values. By default, gaps are filled with `NA` if `x` is atomic, and with `list(NULL)` if `x` is recursive. The user can change the filling value through the `fill_value` argument.

Once again, we can get the original array back when we're done like so:

```{r}
asplit(out, ndim(out)) |> bind_array(along = margin)
```

... but we do keep the missing values when the groups have an unequal number of elements.

&nbsp;

# Nested Binding with Broadcasting

Consider the following 2 lists `x` and `y`:

```{r}
x <- list(
  group1 = list(
    class1 = list(
      height = rnorm(10, 170),
      weight = rnorm(10, 80),
      sex = sample(c("M", "F", NA), 10, TRUE)
    ),
    class2 = list(
      height = rnorm(10, 170),
      weight = rnorm(10, 80),
      sex = sample(c("M", "F", NA), 10, TRUE)
    )
  ),
  group2 = list(
    class1 = list(
      height = rnorm(10, 170),
      weight = rnorm(10, 80),
      sex = sample(c("M", "F", NA), 10, TRUE)
    ),
    class2 = list(
      height = rnorm(10, 170),
      weight = rnorm(10, 80),
      sex = sample(c("M", "F", NA), 10, TRUE)
    )
  )
)


y <- list(
  group1 = list(
    class1 = list(
      smoker = sample(c(TRUE, FALSE), 10, TRUE)
    ),
    class2 = list(
      smoker = sample(c(TRUE, FALSE), 10, TRUE)
    )
  ),
  group2 = list(
    class1 = list(
      smoker = sample(c(TRUE, FALSE), 10, TRUE)
    ),
    class2 = list(
      smoker = sample(c(TRUE, FALSE), 10, TRUE)
    )
  )
)

```

We wish 2 merge these 2 lists, such that after every list-element "sex" in list `x`, the element "smoker" follows in list `y`.

This can be done with a double for-loop in `r fa("r-project")`, but if both lists are large, this becomes slow. With the 'broadcast' package, one can cast both lists as dimensional, and then use broadcasted binding to merge `x` and `y` together with ease. Moreover, if a list can be represented dimensionally, it is advised to make the list dimensional as that will make further analyses, manipulations, and sub-set operations much easier and faster compared to a hierarchical list.

So let's first cast both lists from hierarchical lists to dimensional lists:

```{r}
(x2 <- cast_hier2dim(x, direction.names = 1))
(y2 <- cast_hier2dim(y, direction.names = 1))
```


Now bind these lists together using [bind_array()](/man/bind_array.qmd), which can handle recursive arrays (i.e. dimensional lists) natively with ease:

```{r}
z <- bind_array(list(x2, y2), 1L)
print(z)
```

What we've just done would require an **expensive** double for-loop in base `r fa("r-project")`.
But notice how easily we've achieved this goal with 'broadcast': just 2 steps - casting and then binding!

If you *really*  want `z` to be a nested list, you can cast it back to hierarchical:

```{r}
z2 <- cast_dim2hier(z)
```

Remember that elements of a lists do not contain the actual data; they **point** to the data; casting lists from dimensional to hierarchical and vice-versa only makes inexpensive shallow copies, not deep copies. It's also quite fast.

&nbsp;


# Nested Arithmetic with Broadcasting

We have a class of students who scored certain amounts of points for homework exercises. The software used for collecting the points for these homework exercises produces on object that is read by `r fa("r-project")` as a nested list, like so:

```{r}
x <- list(
  student1 = list(
    homework1 = sample(0:100, 5),
    homework2 = sample(0:100, 5),
    homework3 = sample(0:100, 5)
  ),
  student2 = list(
    homework1 = sample(0:100, 5),
    homework2 = sample(0:100, 5),
    homework3 = sample(0:100, 5)
  ),
  student3 = list(
    homework1 = sample(0:100, 5),
    homework2 = sample(0:100, 5),
    homework3 = sample(0:100, 5)
  )
)
```

Since all values are numbers, the teacher would like the list to be turned in a numeric array, to make analysis easier.

This can be done with the 'broadcast' package with the following steps.

First, turn the nested list into a shallow dimensional list:

```{r}
x2 <- cast_hier2dim(x, in2out = FALSE, direction.names = 1L)
print(x2)
```

Second, turn the dimensional list into an atomic array using [cast_shallow2atomic()](/man/cast_shallow2atomic.qmd):

```{r}
x3 <- cast_shallow2atomic(x2, 1L)
print(x3)
```

Now that we have a numeric array, broadcasted numeric operations can be performed on it. In this case, the teacher would like to multiply the scores of the second homework assignment by 2, as it was deemed much harder than the other 2 homework assignments.  
This can be done like so:

```{r}
# creating multiplier array:
multiplier <- vector2array(c(1L, 2L, 1L), 3L, 3L, TRUE)
dimnames(multiplier)[[3L]] <- dimnames(x3)[[3L]]
print(multiplier)

# broadcasted multiply x3 with multiplier:
broadcaster(x3) <- TRUE
x3 * multiplier
```

Thus we started off with a nested list, through which one cannot broadcast and for which little to no vectorized operators are available.  
By turning the nested list into a dimensional atomic array, both broadcasted and vectorized operations can more easily be used.


&nbsp;

&nbsp;
