---
title: "Installation"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(fontawesome)
```


To install 'broadcast' from GitHub, one may run the following code in `r fa("r-project")`:

```{r echo=TRUE, eval=FALSE}

remotes::install_github("https://github.com/tony-aw/broadcast")

```


'broadcast' is not a large package per-sé, but it does contain a large amount of compiled code ('C' and 'C++'). Compilation of the package may therefore be slow on some operating systems, especially Linux systems; please be patient. On an average PC running Windows 11, installation of the 'broadcast' package shouldn't take more than just a few minutes.

If installation is too slow, one may try to speed up the installation using multi-threading, for example like so:

```{r echo=TRUE, eval=FALSE}

nthreads <- 8L # specify number of threads
Sys.setenv(MAKEFLAGS = paste0("-j", nthreads)) # may speed up installation

remotes::install_github("https://github.com/tony-aw/broadcast")

```


&nbsp;

To attach the package - thus expose its functions to the namespace - please use the following:

```{r echo=TRUE, eval=FALSE}

library(broadcast)

```


&nbsp;

To open the introduction help page of the ‘broadcast’ package, please use any of the following:

```{r echo=TRUE, eval=FALSE}

?broadcast::broadcast
?broadcast::broadcast_help
?broadcast::`broadcast-package`

```


&nbsp;
&nbsp;
