---
title: "Quickstart Guide"
format:
  html:
    number-sections: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(fontawesome)
library(broadcast)
library(abind)
```

# Prerequisites

First, a basic understanding of `r fa("r-project")` is important.  
A very, very basic refresher for `r fa("r-project")` can be found in the [Getting Started in R: Tinyverse Edition](https://eddelbuettel.github.io/gsir-te/Getting-Started-in-R.pdf) document.

&nbsp;

# Installation

To install 'broadcast' from GitHub, one may run the following code in `r fa("r-project")`:

```{r echo=TRUE, eval=FALSE}

remotes::install_github("https://github.com/tony-aw/broadcast")

```


&nbsp;

# Broadcasting

## Introduction

In the context of operations involving 2 (or more) arrays, "broadcasting" refers to recycling array dimensions **without** allocating additional memory, which is considerably **faster** and **more memory-efficient** than R's regular dimensions replication mechanism.

&nbsp;

## Rules

To paraphrase Numpy's own documentation, one can summarise how broadcasting behaves using 2 rules.

 1) if the input arrays for an operation do not have the same number of dimensions, a `1` will be repeatedly appended to the end of the dimension of the smaller array, until the arrays have the same number of dimensions.

2) arrays with a size of 1 for a particular dimension act as if they had the size of the array with the largest size for that dimension. This is done by vritually recycling said dimension without making copies.

After application of these 2 broadcasting rules, the sizes of the input arrays must match.

Please read the [Broadcasting explained](https://tony-aw.github.io/broadcast/vignettes/d_broadcasting_explained.html) page for a more complete explanation of what "broadcasting" is and how it works.

&nbsp;

## Example

Consider the matrices `x` and `y`:

```{r}
x <- array(1:20, c(4, 5))
y <- array(1:5 * 100, c(1, 5))
print(x)
print(y)
```

Suppose one wishes to compute the element-wise addition of these 2 arrays.

This won't work in base `r fa("r-project")`:

```{r eval=FALSE, echo=TRUE}
x + y
Error in x + y : non-conformable arrays
```

You *could* do the following....

```{r}
x + y[rep(1L, 4L),]
```

... but this becomes an issue when `x` and/or `y` become very large, as the above operation involves replicating/copying `y` several times - which costs memory, reduces speed, and the code is not easily scalable for arrays with different dimensions.

The 'broadcast' package performs "broadcasting", which can do the above, but **faster**, **without unnecessary copies**, and scalable to arrays of any size (up to 16 dimensions).  
Like so:

```{r}
bc.d(x, y, "+")
```


&nbsp;

# Binary operations

'broadcast' supports a wide range of binary element-wise broadcasted operations.   
This includes arithmetic, relational, logical, string, and bit-wise operations.

The `bc.*` functions, like [bc.d()](/man/bc.d.qmd) and [bc.b()](/man/bc.b.qmd), provide great control over these operations.

For example:

```{r}
x <- array(1:20, c(4, 5))
y <- array(1:5 * 100, c(1, 5))
print(x)
print(y)

bc.d(x, y, "+")

```

&nbsp;

# Overloaded Operators

Sometimes broadcasting is needed in a large mathematical expression, involving multiple variables, where precedence is of importance.
For example in an expression like this:  

`x + y / z`

Using the `bc.*` functions for that, while possible, may be inconvenient. It may be more convenient to use the base operators directly, whilst still keeping the broadcasting property.  

To this end, the 'broadcast' package provides the [broadcaster](/man/broadcaster.qmd) class, which comes with its own method dispatch for the base operators.

See the following example:

```{r}

x <- array(1:20, c(4, 5))
y <- array(1:5 * 100, c(1, 5))
z <- array(20:1, c(4, 5))

broadcaster(x) <- TRUE
broadcaster(y) <- TRUE
broadcaster(z) <- TRUE

x + y / z

```

&nbsp;

# Array binding

The battle-tested `abind()` function is often used to bind arrays along any arbitrary dimension.

'broadcast' provides an alternative to this, namely the [bind_array()](/man/bind_array.qmd) function, which allows for broadcasting (obviously), and is also notably faster and more memory efficient than `abind()`.


Consider the following arrays:

```{r}

x <- array(1:20, c(4, 5))
y <- array(1:5*10, c(1, 5))
print(x)
print(y)

```

Binding them together with `abind()` won't work:

```{r eval=FALSE, echo=TRUE}
abind::abind(x, y, along = 2)
Error in abind::abind(x, y, along = 2) : 
  arg 'X2' has dims=1, 5; but need dims=4, X
```

To bind `x` and `y` together along columns, `y` needs its single row to be recycled (broadcasted) 4 times.

This can be done in a highly efficient way using [bind_array()](/man/bind_array.qmd), like so:

```{r}
bind_array(list(x, y), 2L)
```


&nbsp;

# Broadcasted General functions

'broadcast' provides the [bcapply()](/man/bcapply.qmd) function, which is a broadcasted apply-like function that applies a function between 2 arrays with broadcasting.

'broadcast' also provides the [bc_ifelse()](/man/bc_ifelse.qmd) function, which is a broadcasted version of `ifelse()`.

&nbsp;

# Other notable functions

The [acast()](/man/acast.qmd) function casts an array into a new dimension.  
Roughly speaking, it is somewhat analogous to `data.table::dcast()`, except that [acast()](/man/acast.qmd) works on arrays (instead of data.tables) and casts into a entirely new dimension (instead of into more columns).


'broadcast' offers type-casting functions. Unlike base R's type-casting functions (`as.logical()`, `as.integer()`, etc.), the type-casting functions from 'broadcast' preserve names and dimensions.

&nbsp;

