---
title: "Quickstart Guide"
format:
  html:
    number-sections: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(fontawesome)
library(broadcast)
library(abind)
```

# Prerequisites

First, a basic understanding of `r fa("r-project")` is important.  
A very, very basic refresher for `r fa("r-project")` can be found in the [Getting Started in R: Tinyverse Edition](https://eddelbuettel.github.io/gsir-te/Getting-Started-in-R.pdf) document.

&nbsp;

# Installation

'broadcast' requires `r fa("r-project")` version 4.2.0 or later and 'Rcpp' version 1.0.14 or later. Windows users also need 'RTools'.

To install & load 'broadcast' from CRAN, one may run the following code in `r fa("r-project")`:

```{r echo=TRUE, eval=FALSE}

install.packages("broadcast", type = "source")
library(broadcast)

```


&nbsp;

# Broadcasting

## Introduction

In the context of operations involving 2 (or more) arrays, "broadcasting" refers to recycling array dimensions **without** allocating additional memory, which is considerably **faster** and **more memory-efficient** than `r fa("r-project")`'s regular dimensions replication mechanism.


&nbsp;

## Example

Consider the matrices `x` and `y`:

```{r}
x <- array(1:20, c(4, 5))
y <- array(1:5 * 100, c(1, 5))
print(x)
print(y)
```

Suppose one wishes to compute the element-wise addition of these 2 arrays.

This won't work in base `r fa("r-project")`:

```{r eval=FALSE, echo=TRUE}
x + y
Error in x + y : non-conformable arrays
```

You *could* do the following....

```{r}
x + y[rep(1L, 4L),]
```

... but if `x` and/or `y` is very large, it will be slow and may even lead to an error:

```{r eval=FALSE, echo=TRUE}
Error: cannot allocate vector of size
```


The 'broadcast' package performs "broadcasting", which can do the above, but **faster**, **without unnecessary copies**, and scalable to arrays of any size (up to 16 dimensions).  

Like so:

```{r}

broadcaster(x) <- TRUE
broadcaster(y) <- TRUE

x + y

```

&nbsp;

## Rules

To paraphrase Numpy's own documentation, one can summarise how broadcasting behaves using 2 rules.

 1) if the input arrays for an operation do not have the same number of dimensions, a `1` will be repeatedly appended to the end of the dimension of the smaller array, until the arrays have the same number of dimensions.

2) arrays with a size of 1 for a particular dimension act as if they had the size of the array with the largest size for that dimension. This is done by virtually recycling said dimension without making copies.

After application of these 2 broadcasting rules, the sizes of the input arrays must match.

&nbsp;

# Broadcasted Infix Operators

Base `r fa("r-project")` comes with relational (==, !=, etc.), arithmetic (+, -, *, /, etc.), logical (&, |) and bit-wise (&, |) operators.
'broadcast' provides 2 ways to use these operators with broadcasting.

The first (and simple) way is to use the [broadcaster()](/man/broadcaster.qmd) class, which comes with it's own method dispatch for the above mentioned operators.
This approach supports operator precedence, and for the average `r fa("r-project")` user, this is sufficient.

For example:

```{r}

x <- array(1:20, c(4, 5))
y <- array(1:5 * 100, c(1, 5))
z <- array(20:1, c(4, 5))

broadcaster(x) <- TRUE
broadcaster(y) <- TRUE
broadcaster(z) <- TRUE

x + y / z

```

The second way is to use the large set of` bc.` - functions.
These offer much greater control and more operators than the previous method, and has less risk of running into conflicting methods.
But it does not support operator precedence.


For example:

```{r}
x <- array(1:20, c(4, 5))
y <- array(1:5 * 100, c(1, 5))
print(x)
print(y)

bc.i(x, y, "gcd") # calculate greatest common divisor between x and y

```


&nbsp;

# Array binding

The battle-tested `abind()` function is often used to bind arrays along any arbitrary dimension.

'broadcast' provides an alternative to this, namely the [bind_array()](/man/bind_array.qmd) function, which allows for broadcasting (obviously), and is also notably faster and more memory efficient than `abind()`.


Consider the following arrays:

```{r}

x <- array(1:20, c(4, 5))
y <- array(1:5*10, c(1, 5))
print(x)
print(y)

```

Binding them together with `abind()` won't work:

```{r eval=FALSE, echo=TRUE}
abind::abind(x, y, along = 2)
Error in abind::abind(x, y, along = 2) : 
  arg 'X2' has dims=1, 5; but need dims=4, X
```

To bind `x` and `y` together along columns, `y` needs its single row to be recycled (broadcasted) 4 times.

This can be done in a highly efficient way using [bind_array()](/man/bind_array.qmd), like so:

```{r}
bind_array(list(x, y), 2L)
```


&nbsp;

# Casting

## Overview

'broadcast' provides several "casting" functions.
These can facility complex forms of broadcasting that would normally not be possible.  
But these "casting" functions also have their own merit, beside empowering complex broadcasting.

The following casting functions are currently available:


 - [acast()](/man/acast.qmd): casts group-based subset of an array into a new dimension.  
Useful for, for example, computing grouped broadcasted operations.

 - [cast_hier2dim()](/man/cast_hier2dim.qmd): casts a nested/hierarchical list into a dimensional list (i.e. recursive array).  
Useful because one cannot broadcast through nesting, but one can broadcast along dimensions.

 - [cast_dim2hier()](/man/cast_dim2hier.qmd): casts a dimensional list into a nested/hierarchical list; the opposite of cast_hier2dim.

 - [cast_dim2flat()](/man/cast_dim2flat.qmd): casts a dimensional list into a flattened list, but with names that indicate their original dimensional positions.  
Mostly useful for printing or summarizing dimensional lists.

 - [dropnests()](/man/dropnests.qmd): drop redundant nesting in lists; mostly used for facilitating the above casting functions.


&nbsp;

## Example

As an example, consider the following nested list:

```{r}
x <- list(
  group1 = list(
    class1 = list(
      height = rnorm(5, 170) |> as.integer(),
      weight = rnorm(5, 80)  |> as.integer(),
      sex = sample(c("M", "F", NA), 5, TRUE)
    ),
    class2 = list(
      height = rnorm(5, 170)  |> as.integer(),
      weight = rnorm(5, 80) |> as.integer(),
      sex = sample(c("M", "F", NA), 5, TRUE)
    )
  ),
  group2 = list(
    class1 = list(
      height = rnorm(5, 170) |> as.integer(),
      weight = rnorm(5, 80) |> as.integer(),
      sex = sample(c("M", "F", NA), 5, TRUE)
    ),
    class2 = list(
      height = rnorm(5, 170) |> as.integer(),
      weight = rnorm(5, 80) |> as.integer(),
      sex = sample(c("M", "F", NA), 5, TRUE)
    )
  )
)

```


One can turn the nested list into a dimensional list using [cast_hier2dim()](/man/cast_hier2dim.qmd) like so:

```{r}
x2 <- cast_hier2dim(x, direction.names = 1L)
print(x2)

```

And turn it back into a dimensional list using [cast_dim2hier()](/man/cast_dim2hier.qmd) like so:

```{r}
x3 <- cast_dim2hier(x2, distr.names = TRUE)
lobstr::tree(x3)
```



&nbsp;


# Other Functions

'broadcast' provides the [bcapply()](/man/bcapply.qmd) function, which is a broadcasted apply-like function that applies a function between 2 arrays with broadcasting.

'broadcast' also provides the [bc_ifelse()](/man/bc_ifelse.qmd) function, which is a broadcasted version of `ifelse()`.

'broadcast' provides a small set of simple linear algebra functions for usage in statistics. See [linear_algebra_stats](/man/linear_algebra_stats.qmd).

'broadcast' offers type-casting functions. Unlike base `r fa("r-project")`'s type-casting functions (`as.logical()`, `as.integer()`, etc.), the type-casting functions from 'broadcast' preserve names and dimensions. See [typecast](/man/typecast.qmd).

&nbsp;


# Why use 'broadcast'?

To keep it very short:

 - using broadcasting is several times faster and uses several times less memory than using base `r fa("r-project")` alternative approaches like using `sweep()`, `outer()`, or manual replication of dimensions.  
 See the [benchmarks](/about/f_benchmarks_other.qmd).
 - 'broadcast' provides [bind_array()](/man/bind_array.qmd), which is an `abind()`-like function that can bind arrays along any arbitrary dimension.  
 The difference with `abind()` is that [bind_array()](/man/bind_array.qmd) supports broadcasting, supports recursive arrays, and is faster & more memory-efficient.
 - 'broadcast' can perform (admittedly niche but still useful) forms of manipulations on arrays and hierarchical lists not found (to my knowledge) in other packages.

&nbsp;

&nbsp;
