---
title: "Broadcasting Explained"
format:
  html:
    number-sections: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, eval=TRUE, echo=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
library(broadcast) |> suppressMessages()
library(fontawesome) |> suppressWarnings()
library(tinytable) |> suppressWarnings()
```


&nbsp;

# Introduction

In the context of operations involving 2 (or more) arrays, “broadcasting” refers to efficiently recycling array dimensions, without making copies.  
This is considerably **faster** and **more memory-efficient** than `r fa("r-project")`'s regular dimensions replication mechanism.

This page explains the concept of "broadcasting" in more detail.
A good understanding of atomic and recursive arrays in base `r fa("r-project")` is somewhat essential to follow this page.

&nbsp;

# What is broadcasting and why is it needed?

## Example case

Let's start with a simple example.  
Consider the matrices `x` and `y`:

```{r}
x <- array(1:20, c(4, 5))
y <- array(1:5*10, c(1, 5))
print(x)
print(y)
```

Suppose one wishes to compute the element-wise addition of these 2 arrays.  

This won't work in base `r fa("r-project")`:

```{r eval=FALSE, echo=TRUE}
x + y
Error in x + y : non-conformable arrays
```


When computing the element-wise sum of these arrays, one or both of them need to be recycled so that they are equal size, in order to compute the element-wise computation.

In this case, matrix `y` needs its single row to be recycled 4 times, making   `y` the same size as `x`, and thus conformable.  

`r fa("r-project")` provides linear **vector** recycling, but not recycling of array dimensions. Instead, in base `r fa("r-project")` we need to **replicate** (and thus **copy**) the array dimensions. This can be done manually, like so:

```{r}
x + y[rep(1L, 4L),]
```


&nbsp;

## What is the problem?

There are several problems.

First and foremost, when the arrays become larger, the replicated arrays become larger, and if the arrays become too large, you may require more memory than is available in your current system, resulting in a message like the following:

```{r eval=FALSE, echo=TRUE}
Error: cannot allocate vector of size
```


The problem isn't limited to available memory. As the required memory to allocate increase, the speed decreases. And let's not forget that computational inefficiency in general is also bad for the environment.

Additionally, a solution like `x + y[rep(1L, 4L),]` is not easily scalable for other arrays when the dimensions of `x` and `y` are not known a-priori.

&nbsp;

## Introducing Broadcasting

In an operation like `x + y[rep(1L, 4L),]`, `y` is replicated to become the same size as `x`. The thing is, physical replication - and thus copying - of the dimensions of an array should not be necessary; arrays only need to be recycled virtually.  

Virtual recycling does not actually physically replicate arrays. Instead, nested loops in 'C' and 'C++' are used to simulate a recycled array.  
This is similar to how `r fa("r-project")` recycles scalars to vectors.  
Virtual recycling requires no additional memory (apart from allocating the final end result); it is much faster and much more memory efficient than replicating dimensions.

And **that** is what broadcasting does: broadcasting provides **fast** virtual recycling of array dimensions in the context of element-wise operations involving 2 (or more) arrays **without** allocating additional memory. Broadcasting in this package is also scalable to arrays of any dimensions (up to 16 dimensions).

In the earlier example, we used:

```{r}
x + y[rep(1L, 4L),]
```

To compute the element-wise addition using broadcasting through the 'broadcast' package, one can mark both `x` and `y` as "broadcaster", and then use the `+` operator, like so:

```{r}

broadcaster(x) <- TRUE
broadcaster(y) <- TRUE

x + y

```

The result is the same (as it should). But as the size of the resulting array increases, the broadcasted functions become more and more efficient in terms of both speed and memory, in comparison with base `r fa("r-project")` approaches.  

Benchmarks can be found on the website.

&nbsp;

# Illustrating Broadcasting

## Matrices

Consider the following matrices

```{r}
x <- array(1:12, c(4, 3))
y <- array(1:5*10, c(1, 3))
broadcaster(x) <- TRUE
broadcaster(y) <- TRUE
print(x)
print(y)
```

Let's see what happens if we sum them together using broadcasting:

```{r}

x + y

```

What happens here is that `x` remains the same, but each row of `y` is **virtually** recycled 4 times, but without requiring 4 times more memory. This is also illustrated here:

```{r eval=TRUE, echo=FALSE}


foo <- rbind(
  sprintf("x[%d, %d] + y[1, %d]", 1, 1:3, 1:3),
  sprintf("x[%d, %d] + y[1, %d]", 2, 1:3, 1:3),
  sprintf("x[%d, %d] + y[1, %d]", 3, 1:3, 1:3),
  sprintf("x[%d, %d] + y[1, %d]", 4, 1:3, 1:3)
)

foo <- as.data.frame(foo)
colnames(foo) <- NULL
tt(foo, theme = "grid") |> style_tt(line = "tlbr")


```

&nbsp;

## Perpendicular vectors

Let's now consider another scenario.
We now take again 2 matrices `x` and `y`, but this time `x` is a matrix with 1 column and multiple rows, and `y` is a matrix with 1 row and multiple columns:


```{r}
x <- array(1:3, c(3, 1))
y <- array(1:3 * 10, c(1, 3))
broadcaster(x) <- TRUE
broadcaster(y) <- TRUE
print(x)
print(y)
```


Computing the broadcasted element-wise sum of `x` and `y` produces the following:

```{r}

x + y

```


So what exactly does broadcasting compute here?

The following table illustrates what broadcasting does, conceptually:

```{r eval=TRUE, echo=FALSE}


foo <- cbind(
  sprintf("x[%d, 1] + y[1, %d]", 1:3, 1),
  sprintf("x[%d, 1] + y[1, %d]", 1:3, 2),
  sprintf("x[%d, 1] + y[1, %d]", 1:3, 3)
)
foo <- as.data.frame(foo)
colnames(foo) <- NULL
tt(foo, theme = "grid") |> style_tt(line = "tlbr")


```

&nbsp;


# Vector Orientation & Length

In the documentation of 'broadcast', the reader may occasionally find references to the "orientation" of a vector. The orientation of a vector can mean several different things, depending on the context. Before going further, we need to address what the orientation of a vector means in the context of broadcasting.

Let's consider the following array `z`:

```{r}
z <- array(1:10, c(1, 10))
print(z)
```

Here, `z` has 1 row and 10 columns.
It is an array, but effectively it is also a vector: it has only one dimension with size larger than 1.

An oriented vector is an array where only one dimension is larger than 1.
The orientation of a vector refers to **which** dimension is larger than 1.

`z` is an oriented vector, and its orientation is 2.

Now the most important part:  
Following the standard rules of Linear Algebra, a vector `z` without dimensions is interpreted by the 'broadcast' package as a vector with orientation `1`.
For example, the vectors `x` and `y` below are treated equally by the 'broadcast' package:

```{r}
x <- array(1:10, c(10, 1))
y <- 1:10
print(x)
print(y)
```


&nbsp;

Since dimensionless vectors are interpreted as oriented vectors (which is necessary to meaningfully perform broadcasting), when may wonder: how does the 'broadcast' package deal with  `long` vectors?

The answer is simple.  
Given some broadcasted operation involving arrays or vectors `x` and `y` (for example `x + y`), the following holds.  
If the result will be a vector, long vectors are supported. I.e. if `x` and `y` are both vectors, than `x` and/or `y` may be long vectors.  
However, if, for example, `x` is a long vector, and `y` is an array (regardless if it is an oriented vector or not), an error is returned: this is because arrays in `r fa("r-project")` cannot have dimensions larger than `2^31 - 1`.

::: {.callout-tip}

One can easily change the orientation of an oriented vector using the new `%orientbc<-%` operator (available from version 0.1.9.5).

:::


&nbsp;

# Broadcasting limitations and rules

The examples shown so far used matrices. But 'broadcast' is not restricted to only handling matrices; this package can handle any array, provided it does not have more than 16 dimensions.
Moreover, any number of dimensions (provided <= 16) can be broadcasted simultaneously.

The examples show so far only used addition. But 'broadcast' supports a wide variety of binary operations, including relational-, arithmetic-, Boolean-, string- and bit-wise operations.


&nbsp;

## Normalization

Before broadcasting occurs, the number of dimensions of `x` and `y` are normalized, such that `ndim(x)` and `ndim(y)` are the same (`ndim(x)` is the same as `length(dim(x))`).

This normalization is achieved by appending dimensions of size 1 to the array with the smaller number of dimensions until both array have the same number of dimensions.

For example, if `dim(x) = c(3, 4, 3)` and `dim(y) = c(3, 4)`, then internally `dim(y)` is *interpreted* to be `c(3, 4, 1)`. This does not actually copy or even modify array `y`.

Following the standard rules of Linear Algebra, a vector `z` without dimensions is taken to be an oriented vector with orientation 1.

&nbsp;


## Conformability

Broadcasted operations only work when the arrays are conformable for broadcasting.  

Consider again 2 arrays `x` and `y`, and their dimensions given by `dim(x)` and `dim(y)`, respectively.  
'broadcast' goes through `dim(x)` and `dim(y)`, from left (i.e. first dimension / rows) to right (i.e. the last dimension), and checks for each axis `i` if **at least one** of the following conditions is TRUE:

 - `dim(x)[i]` and `dim(y)[i]` are equal;
 - either `dim(x)[i]` or `dim(y)[i]` is `1`;
 - either `dim(x)[i]` or `dim(y)[i]` is non-existing (see also the "Normalization" sub-section above)
 
if at least one of the above conditions is true for each and every one of the dimensions of `x` and `y`, the 2 arrays are compatible for broadcasted operations. If not, they are not compatible, and attempting to perform a broadcasted operations results in an error.


To illustrate, let's check 2 arrays to see if they are compatible:

```{r echo = TRUE, eval = TRUE}
x <- array(rnorm(10), c(10, 1, 9, 6))
y <- array(rnorm(10), c(10, 5, 1))
```


```{r echo=FALSE, eval=TRUE}
dimx <- dim(x)
dimy <- dim(y)
foo <- data.frame(
  i = 1:4,
  `dim(x)[i]` = dimx,
  `dim(y)[i]` = c(dimy, NA),
  compatible = rep(TRUE, 4),
  reason = c("equal", "either is 1", "either is 1", "either is missing"),
  check.names = FALSE
)

tt(foo)

```

As shown in the data.frame above, all dimensions of `x` and `y` are compatible.  
Therefore, broadcasted binary operations involving `x` and `y` can be performed.

&nbsp;

Now let's look at another 2 arrays, and see if they are compatible:


```{r echo = TRUE, eval=TRUE}
x <- array(rnorm(10), c(10, 1, 9))
y <- array(rnorm(10), c(10, 5, 2, 6))
```


```{r echo=FALSE, eval=TRUE}
dimx <- dim(x)
dimy <- dim(y)
foo <- data.frame(
  i = 1:4,
  `dim(x)[i]` = c(dimx, NA),
  `dim(y)[i]` = dimy,
  compatible = c(TRUE, TRUE, FALSE, TRUE),
  reason = c("equal", "either is 1", "not equal & neither is 1/missing", "either is missing"),
  check.names = FALSE
)

tt(foo)

```

One of the dimensions, namely the third dimension, is not compatible. Therefore, these 2 arrays are **not** conformable for broadcasted operations.

&nbsp;

## Which dimensions are broadcasted

In the context of an operation involving exactly 2 arrays, this sub-section explains which dimensions of which array is broadcasted, and which dimension of which array is left as-is.  
There are 3 scenarios relevant for this explanation.

&nbsp;

**Scenario 1:**  
For some dimension `i`, `dim(x)[i]` and `dim(y)[i]` are equal.  
In this scenario, no broadcasting needs to occur on that dimension for either array.

&nbsp;


**Scenario 2:**  
For some dimension `i`, suppose `dim(x)[i] == 1` and `dim(y)[i] > 1`.  
Then dimension `dim(x)[i]` is **broadcasted** to size `dim(y)[i]`.  
Similarly, if `dim(x)[i] > 1` and `dim(y)[i] == 1`, dimension `dim(y)[i]` is **broadcasted** to size `dim(x)[i]`.

&nbsp;


**Scenario 3:**  
Finally, suppose for some dimension `i`, `dim(x)[i] > 1`, and `dim(y)[i]` is non-existing.  
I.e., `x` has more dimensions than `y`.  
In this case, the missing dimension in `y` is replaced with a dimensions of size `1`, and scenario 2 is used.  
Similarly, if `dim(y)[i] > 1` and `dim(x)[i]` is non-existing, the missing dimension in `x` is replaced with a dimension of size `1` and we get back to scenario 2.  

It does not matter how much difference there is between the number of dimensions of `x` and the number of dimensions of `y`: any number of missing dimensions will be replaced with `1`, as long as the total number of dimensions for each array does not exceed `16`.

&nbsp;

**Illustration**

To illustrate, let's check 2 conformable arrays, and see which dimension of which array is broadcasted by how much

```{r echo = TRUE, eval = TRUE}
x <- array(rnorm(10), c(10, 1, 1, 9, 6))
y <- array(rnorm(10), c(10, 1, 5, 1))
```


```{r echo=FALSE, eval=TRUE}
dimx <- dim(x)
dimy <- dim(y)
foo <- data.frame(
  i = 1:length(dimx),
  `dim(x)[i]` = dimx,
  `dim(y)[i]` = c(dimy, NA),
  broadcasted = c("neither", "neither", "x", "y", "y"),
  reason = c(
    "dims equal", "dims equal",
    "dim(x)[i] == 1", "dim(y)[i] == 1",
    "dim(y)[i] is missing"
  ),
  note = c(
    rep("", 2),
    "dim(x)[i] recycled to size 5",
    "dim(y)[i] recycled to size 9",
    "dim(y) extended with a 1; dim(y)[i] recycled to size 6"
  ),
  check.names = FALSE
)

tt(foo)

```


&nbsp;

# Orthogonal Arrays

In the documentation of 'broadcast', the reader may come across the term "orthogonal arrays" here and there.  
"orthogonal" is a term that can mean a great many things; this section therefore explains what "orthogonal" means in the context of broadcasting.

Consider the oriented vectors `x` and `y`, where their orientations are `1` and `2`, respectively:

```{r}
x <- array(1:5, c(5, 1))
y <- array(1:5*10, c(1, 5))
print(x)
print(y)
```

These vectors are (in a very literal sense) perpendicular to each other, and thus in that sense orthogonal to each other.  
This type of orthogonality can be extended to arrays as follows.  
For the purposes of broadcasting, any pair of arrays, `x` and `y`, can be referred to as being "orthogonal" if **all** of the following holds for that pair of arrays:  

 - For **every** dimension index `i`, `dim(x)[i] != dim(y)[i]`
 - For **every** dimensions index `i`,`dim(x)[i]` is `1` or missing, **OR** `dim(y)[i]` is `1` or missing.


&nbsp;

## Illustration


To illustrate, here are 2 arrays that are fully orthogonal:

```{r echo = TRUE, eval = TRUE}
x <- array(1:10, c(10, 1, 8, 1, 10))
y <- array(1:10, c(1, 9, 1, 9, 1, 11))
```


```{r echo=FALSE, eval=TRUE}
dimx <- dim(x)
dimy <- dim(y)
foo <- data.frame(
  i = 1:6,
  `dim(x)[i]` = c(dimx, NA),
  `dim(y)[i]` = dimy,
  orthogonal = rep(TRUE, 6L),
  reason = c(rep("dim(x)[i] != dim(y)[i]", 5), "one of them missing"),
  check.names = FALSE
)

tt(foo)

```

&nbsp;

And the following 2 arrays are **not** orthogonal:


```{r echo=TRUE, eval=TRUE}
x <- array(1:10, c(10, 1, 8, 1, 10))
y <- array(1:10, c(10, 9, 1, 9, 10))

```

```{r echo=FALSE, eval=TRUE}
dimx <- dim(x)
dimy <- dim(y)
foo <- data.frame(
  i = 1:5,
  `dim(x)[i]` = dimx,
  `dim(y)[i]` = dimy,
  orthogonal = c(FALSE, rep(TRUE, 3L), FALSE),
  reason = c("dim(x)[i] == dim(y)[i]", rep("dim(x)[i] != dim(y)[i]", 3), "dim(x)[i] == dim(y)[i]"),
  check.names = FALSE
)

tt(foo)

```

The above arrays are **not** orthogonal, due to dimensions 1 and 5.


&nbsp;


# Broadcasting vs outer

The base `r fa("r-project")` `outer()` function performs outer computations on vectors and arrays.

Earlier in this guide, we used the following computation:

```{r}
x <- array(1:5, c(5, 1))
y <- array(1:5*10, c(1, 5))
broadcaster(x) <- TRUE
broadcaster(y) <- TRUE

x + y

```


This is equivalent to using the `outer()` function as follows:

```{r}
outer(as.vector(x), as.vector(y), "+")
```


As you can see, performing a broadcasted operation between vectors with orientations 1 and 2 (in that order) is equivalent to using `outer()` on 2 dimensionless vectors.

The `outer()` approach has the same problems as the `x[, rep(1L, 5L)] + y[rep(1L, 5L),]` approach: it consumes a lot of unnecessary memory, and is slow.

Broadcasting can thus be used as a more efficient alternative to `outer()` in case of vector inputs, simply by making the inputs orthogonal oriented vectors.

&nbsp;

# Broadcasted Binding


The [bind_array()](/man/bind_array.qmd) comes with the `ndim2bc` argument (an abbreviation of "maximum number of dimensions to broadcast"), which allows users to specify the maximum number of dimensions that are allowed to be broadcasted while binding arrays. This way, users won't get unpleasant surprises.

This should be fairly obvious, but the dimension specified in `along` is never broadcasted.

&nbsp;


Consider the following arrays:

```{r}

x <- array(1:20, c(4, 5))
y <- array(1:5*10, c(1, 5))
print(x)
print(y)

```

Binding them together with `abind()` won't work:

```{r eval=FALSE, echo=TRUE}
abind::abind(x, y, along = 2)
Error in abind::abind(x, y, along = 2) : 
  arg 'X2' has dims=1, 5; but need dims=4, X
```

To bind `x` and `y` together along columns, `y` needs its single row to be recycled (broadcasted) 4 times.

This can be done in a highly efficient way using [bind_array()](/man/bind_array.qmd), like so:

```{r}
bind_array(list(x, y), 2L)
```

&nbsp;



But what if broadcasting is explicitly **not** desired? What if one actually wants this to produce an error, like `abind()`? Fret not, for that's what the `ndim2bc` argument is for. Setting it to `0` will disable broadcasting altogether:

```{r eval=FALSE, echo=TRUE}
bind_array(list(x, y), 2L, ndim2bc = 0)
Error in bind_array(list(x, y), 2L, ndim2bc = 0) : 
  maximum number of dimensions to be broadcasted (1) exceeds `ndim2bc` (0)
```


&nbsp;


Let's replace `x` with a 3 dimensional array:

```{r}
x <- array(1:20, c(4, 5, 3))
print(x)
```

Suppose we don't want to broadcast more than 1 dimension at a time, out of fear of accidentally broadcasting too many dimensions (like when binding arrays in a loop without knowing a-priori how many dimensions the arrays have).

One can than set `ndim2bc = 1L`, to ensure no more than one dimensions is being broadcasted when binding arrays.

So trying to bind `x` with `y` now will produce an error even with [bind_array()](/man/bind_array.qmd), to protect the user from unintended broadcasting:

```{r eval=FALSE, echo=TRUE}
bind_array(list(x, y), 2L, ndim2bc = 1L)
Error in bind_array(list(x, y), 2L, ndim2bc = 1L) : 
  maximum number of dimensions to be broadcasted (2) exceeds `ndim2bc` (1)
```

&nbsp;


