---
title: "Prerequisites & Installation"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(fontawesome)
```

# Prerequisites

First, a basic understanding of `r fa("r-project")` is important.  
A very, very basic refresher for `r fa("r-project")` can be found in the [Getting Started in R: Tinyverse Edition](https://eddelbuettel.github.io/gsir-te/Getting-Started-in-R.pdf) document.

&nbsp;

Make sure you at least have basic knowledge of the following subjects in `r fa("r-project")`:

 - Atomic types (logical, integer, double, complex, character, raw).
 - The recursive type "list".
 - dimensional structures: vectors, matrices, arrays.
 - common arithmetic operators (+, -, *, /, etc.), relational operators (==, !=, etc.), logical operators (&, |), and common bit-wise operations.
 - vector recycling



&nbsp;

# Installation

To install 'broadcast' from GitHub, one may run the following code in `r fa("r-project")`:

```{r echo=TRUE, eval=FALSE}

remotes::install_github("https://github.com/tony-aw/broadcast")

```


'broadcast' is not a large package per-sé, but it does contain a large amount of compiled code ('C' and 'C++'). Compilation of the package may therefore be slow on some operating systems, especially Linux systems; please be patient. On an average PC running Windows 11, installation of the 'broadcast' package shouldn't take more than just a few minutes.

If installation is too slow, one may try to speed up the installation using multi-threading, for example like so:

```{r echo=TRUE, eval=FALSE}

nthreads <- 8L # specify number of threads
Sys.setenv(MAKEFLAGS = paste0("-j", nthreads)) # may speed up installation

remotes::install_github("https://github.com/tony-aw/broadcast")

```


&nbsp;

To attach the package - thus expose its functions to the namespace - please use the following:

```{r echo=TRUE, eval=FALSE}

library(broadcast)

```


&nbsp;

To open the introduction help page of the ‘broadcast’ package, please use any of the following:

```{r echo=TRUE, eval=FALSE}

?broadcast::broadcast
?broadcast::broadcast_help
?broadcast::`broadcast-package`

```


&nbsp;
&nbsp;
