---
title: "Operator Overloading"
format:
  html:
    number-sections: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo=FALSE, eval=TRUE}
library(broadcast) |> suppressMessages()
library(fontawesome) |> suppressWarnings()
```


&nbsp;

# Introduction

Base 'R' comes with relational (==, !=, etc.), arithmetic (+, -, *, /, etc.), logical (&, |) and bit-wise (&, |) operators.
'broadcast' provides 2 ways to use these operators with broadcasting.

The first (and simple) way is to use the [broadcaster()](/man/broadcaster.qmd) class, which comes with it's own method dispatch for the above mentioned operators.
This approach supports operator precedence, and for the average 'R' user, this is sufficient.

For example:

```{r}

x <- array(1:20, c(4, 5))
y <- array(1:5 * 100, c(1, 5))
z <- array(20:1, c(4, 5))

broadcaster(x) <- TRUE
broadcaster(y) <- TRUE
broadcaster(z) <- TRUE

x + y / z

```

The second way is to use the large set of` bc.` - functions.
These offer much greater control and more operators than the previous method, and has less risk of running into conflicting methods.
But it does not support operator precedence.


For example:

```{r}
x <- array(1:20, c(4, 5))
y <- array(1:5 * 100, c(1, 5))
print(x)
print(y)

bc.i(x, y, "gcd") # calculate the greatest common divisor between x and y

```



&nbsp;

# Derived operators

There is no overload for `xor()`, as `xor()` is defined using the existing base relational and logical (or bit-wise, for type of `raw`) operators.  

There is also no overload for the %d==%, %d!=%, etc. operators, as they too are defined using the existing base relational and logical operators.


&nbsp;

# Overloaded operators vs `bc.*` functions

Overloading the operators is primarily useful for ensuring mathematically correct operator precedence, and to reduce the amount of typing.  
However, the many `bc.*` functions provide much greater control than the simple operators can provide.  

For example:  
The `&` and `|` operators provide logical AND and OR, respectively for all atomic types except type `raw`; for type `raw`, the `&` and `|` operators provide BIT-WISE AND and OR, respectively.  
But what if you want to use logical AND/OR for `raw` arrays, or bit-wise AND/OR for `integer` arrays?  
When using the overloaded operators, using logical AND/OR for type `raw` necessitates converting the vector/array to type `logical`, thus making a copy.  
With the `bc.*()` functions, this is not necessary.  
Want to use logical AND/OR? Use [bc.b()](/man/bc.b.qmd), it supports several types including `raw`.  
Want to use bit-wise AND/OR? Use [bc.bit()](/man/bc.bit.qmd), it supports not only `raw`, but `integer` as well.

The `bc.` functions also provides far more operations than the simple symbols can provide.  
Like calculating the greatest common divisors between integers, string edit distance, and more.

&nbsp;


