---
title: "Functions Overview"
format:
  html:
    number-sections: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, eval=TRUE, echo=FALSE}
library(fontawesome) |> suppressWarnings()
library(broadcast) |> suppressMessages()
```


&nbsp;

# Introduction

The 'broadcast' package provides 6 main functionalities:

 - Broadcasted Infix Operators
 - broadcasted dimensional binding of arrays
 - casting functions, that cast subset-groups of an array to a new dimension, cast nested lists to dimensional lists, and vice-versa.
 - broadcasted generic pair-wise functions
 - array typecasting with dimensions and names preserved
 - simple linear algebra functions for statistics
 
This guide gives a brief overview of the functions provided by the 'broadcast' `r fa("r-project")` - package.


&nbsp;

# Broadcasted Infix Operators

Base `r fa("r-project")` comes with relational (==, !=, etc.), arithmetic (+, -, *, /, etc.), logical (&, |) and bit-wise (&, |) operators.
'broadcast' provides 2 ways to use these operators with broadcasting.

The first (and simple) way is to mark each array involved in the operation as a [broadcaster()](/man/broadcaster.qmd), which comes with it's own method dispatch for the above mentioned operators.
This approach supports operator precedence, and for the average `r fa("r-project")` user, this is sufficient.

For example:

```{r}

x <- array(1:20, c(4, 5))
y <- array(1:5 * 100, c(1, 5))
z <- array(20:1, c(4, 5))

broadcaster(x) <- TRUE
broadcaster(y) <- TRUE
broadcaster(z) <- TRUE

x + y / z

```

The second way is to use the large set of` bc.` - functions.
These offer much greater control and more operators than the previous method, and has less risk of running into conflicting methods.
But it does not support operator precedence.


For example:

```{r}
x <- array(1:20, c(4, 5))
y <- array(1:5 * 100, c(1, 5))
print(x)
print(y)

bc.i(x, y, "gcd") # calculate greatest common divisor between x and y

```



&nbsp;

## Derived operators

There is no overload for `xor()`, as `xor()` is defined using the existing base relational and logical (or bit-wise, for type of `raw`) operators.  

There is also no overload for the %d==%, %d!=%, etc. operators, as they too are defined using the existing base relational and logical operators.


&nbsp;

## Overloaded operators vs `bc.*` functions

Overloading the operators is primarily useful for ensuring mathematically correct operator precedence, and to reduce the amount of typing.  
However, the many `bc.*` functions provide much greater control than the simple operators can provide.  

For example:  
The `&` and `|` operators provide logical AND and OR, respectively for all atomic types except type `raw`; for type `raw`, the `&` and `|` operators provide BIT-WISE AND and OR, respectively.  
But what if you want to use logical AND/OR for `raw` arrays, or bit-wise AND/OR for `integer` arrays?  
When using the overloaded operators, using logical AND/OR for type `raw` necessitates converting the vector/array to type `logical`, thus making a copy.  
With the `bc.*()` functions, this is not necessary.  
Want to use logical AND/OR? Use [bc.b()](/man/bc.b.qmd), it supports several types including `raw`.  
Want to use bit-wise AND/OR? Use [bc.bit()](/man/bc.bit.qmd), it supports not only `raw`, but `integer` as well.

The `bc.` functions also provides far more operations than the simple symbols can provide.  
Like calculating the greatest common divisors between integers, string edit distance, and more.



&nbsp;

# Binding arrays

'broadcast' provides the [bind_array()](/man/bind_array.qmd) function. This binds arrays (including matrices) along any arbitrary dimension. It allows for dimensional broadcasting, and returns an array.


&nbsp;


## Why another array binding function

There is already a famous, battle-tested function for binding arrays, namely the `abind()` function from the fantastic `r fa("r-project")` package of the same name. So why does 'broadcast' provide another array binding function?

Well, as great as `abind::abind()` is, there are some issues with it, and the [bind_array()](/man/bind_array.qmd) function from 'broadcast` attempts (and in my humble opinion succeeds) to provide a better alternative:

 - [bind_array()](/man/bind_array.qmd) is faster and more memory efficient.
 - `abind()` does not support broadcasting. [bind_array()](/man/bind_array.qmd) does support broadcasting.
 - `abind()` does not support recursive arrays (i.e. arrays of type `list`). When binding recursive arrays, `abind()` will attempt to unlist it, even when it does not make sense, ruining the structure of the array. [bind_array()](/man/bind_array.qmd) fully supports recursive arrays, as well as arrays of all atomic types.
 - `abind()` has somewhat confusing mechanisms for giving dimension names to the output. [bind_array()](/man/bind_array.qmd) has more streamlined naming arguments.


[bind_array()](/man/bind_array.qmd) uses a similar API to that of `abind::abind()`, and provides similar features like binding arrays before the first dimension (using `along = 0`) or after the last dimension.


&nbsp;

# Casting

'broadcast' provides several "casting" functions.
These can facilitate complex forms of broadcasting that would normally not be possible.  
But these "casting" functions also have their own merit, beside empowering complex broadcasting.

The following casting functions are currently available:


 - [acast()](/man/acast.qmd): casts group-based subset of an array into a new dimension.  
Useful for, for example, computing grouped broadcasted operations.

 - [cast_hier2dim()](/man/cast_hier2dim.qmd): casts a nested/hierarchical list into a dimensional list (i.e. recursive array).  
Useful because one cannot broadcast through nesting, but one can broadcast along dimensions.

 - [cast_dim2hier()](/man/cast_dim2hier.qmd): casts a dimensional list into a nested/hierarchical list; the opposite of cast_hier2dim.

 - [cast_shallow2atomic()](/man/cast_shallow2atomic.qmd): casts a (dimensional) shallow (i.e. non-nested) list into an atomic vector or array.  
Useful because atomic vectors/arrays have access to many vectorized (broadcasted) operations that may not be available for vectors/arrays of type list.

 - [cast_dim2flat()](/man/cast_dim2flat.qmd): casts a dimensional list into a flattened list, but with names that indicate their original dimensional positions.  
Mostly useful for printing or summarizing dimensional lists.

 - [dropnests()](/man/dropnests.qmd): drop redundant nesting in lists; mostly used for facilitating the above casting functions.
 
 - The `%orientbc<-%` operator: changes the orientation of an oriented vector.
Oriented vectors can be used to modify an array along a dimension.


&nbsp;

# General Broadcasted methods

'broadcast' provides the [bcapply()](/man/bcapply.qmd) method, which is a broadcasted apply-like function.  

'broadcast' also provides the [bc_ifelse()](/man/bc_ifelse.qmd) and [bc_strrep()](/man/bc_strrep.qmd)methods, which are broadcasted versions of `ifelse()` and `strrep()`, respectively.

&nbsp;

# Type-casting

'broadcast' offers type-casting functions. Unlike base R's type-casting functions (`as.logical()`, `as.integer()`, etc.), the type-casting functions from 'broadcast' preserve names and dimensions. See [typecast](/man/typecast.qmd).

&nbsp;

# Simple linear algebra functions for statistics

'broadcast' provides a small set of simple linear algebra functions for usage in statistics. See [linear_algebra_stats](/man/linear_algebra_stats.qmd).

&nbsp;

# Supported Structures

The functions in the 'broadcast' package only support S3 structures. S4 structures, Reference Classes, and so on are not supported.


&nbsp;

# Writing Methods for Custom Classes

Almost all functions provided by 'broadcast' are S3 or S4 generics.  
Users and package authors can write methods for these generics to support custom classes (like custom handling of class-related attributes).  

&nbsp;
