---
title: "Operator Overloading"
format:
  html:
    number-sections: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo=FALSE, eval=TRUE}
library(broadcast) |> suppressMessages()
library(fontawesome) |> suppressWarnings()
```


&nbsp;

# Introduction

Sometimes broadcasting is needed in a large mathematical expression, involving multiple variables, where precedence is of importance.
For example in an expression like this:  

`x / (y + z) + a`

Using the `bc.*` functions for that, while possible, may be inconvenient. It may be more convenient to use the base operators directly, whilst still keeping the broadcasting property.  

To this end, the 'broadcast' package provides the [broadcaster](/man/broadcaster.qmd) class, which comes with its own method dispatch for the base operators.

&nbsp;

# Example

Consider the matrices `x`, `y` and `z`:

```{r}
x <- array(1:20, c(4, 5))
y <- array(1:5 * 100, c(1, 5))
z <- array(20:1, c(4, 5))
print(x)
print(y)
print(z)
```

Suppose we wish to compute `x + y / z`;

The following is **wrong**:

```{r}
bc.d(x, y, "+") |> bc.d(y, "/")
```

...because division must come **before** addition, according to standard math precedence rules.

We *could* do the following:

```{r}
bc.d(y, z, "/") |> bc.d(x, "+")
```

...but as the mathematical expression involves more and more variables with a greater variety of operators, it comes more and more taxing to keep track of the operator precedence.

Instead, one can designate all arrays as a "broadcaster"; all base operators that involve a broadcaster will use the overloaded variant provided by the 'broadcast' package - like so:

```{r}
broadcaster(x) <- TRUE
broadcaster(y) <- TRUE
broadcaster(z) <- TRUE
x + y / z

```


Actually, not *all* arrays necessarily need to be a broadcaster.  
Overloaded arithmetic operations that involve a broadcaster, will result in a broadcaster.  
So technically, one could just make the array that is used first - according to the mathematical rules of precedence - as a broadcaster, and broadcasting will continue to be used.  
In this case, the mathematically first preceding array is `y`.  
To demonstrate, we'll un-set `x` and `z` as broadcaster, and you'll see broadcasting still occurs:

```{r}

broadcaster(x) <- FALSE
broadcaster(z) <- FALSE

x + y / z
```


&nbsp;

# Overload method tables

The overloaded operators from the 'broadcast' package attempts to mimic the base `r fa("r-project")` operators accurately, except that broadcasting is used (obviously).

The following tables overview the behaviour of the operators.

&nbsp;

## Regular Arithmetic operators

If both sides of the operators given here are numeric or logical, the following holds:

```{r echo=FALSE, eval=TRUE}

ops <- c("+", "-", "*", "/", "^", "%/%", "%%")
action <- c("add", "substract", "multiply", "divide", "raise to power", "floored integer divide", "modulo")
bc <- c(rep("bc.d", 5), "bc.i", "bc.i")

df <- data.frame(
  operator = ops,
  action = action,
  `function call` = bc,
  check.names = FALSE
)

tinytable::tt(df)

```

&nbsp;

## Complex Arithmetic operators

If at least one of the arguments of the operators given here are complex, the following holds:


```{r echo=FALSE, eval=TRUE}

ops <- c("+", "-", "*", "/")
action <- c("add", "substract", "multiply", "divide")
bc <- c(rep("bc.cplx", 4))

df <- data.frame(
  operator = ops,
  action = action,
  `function call` = bc,
  check.names = FALSE
)

tinytable::tt(df)

```

&nbsp;

## Logical Operators

If both sides of the operators given here are complex, numeric or logical, the following holds:

```{r echo=FALSE, eval=TRUE}

ops <- c("&", "|")
action <- c("and", "or")
bc <- c(rep("bc.b", 2))

df <- data.frame(
  operator = ops,
  action = action,
  `function call` = bc,
  check.names = FALSE
)

tinytable::tt(df)
```

&nbsp;

## Bit-wise operators


If both sides of the operators given here are type of raw, the following holds:


```{r echo=FALSE, eval=TRUE}

ops <- c("&", "|")
action <- c("bit-wise and", "bit-wise or")
bc <- c(rep("bc.bit", 2))

df <- data.frame(
  operator = ops,
  action = action,
  `function call` = bc,
  check.names = FALSE
)

tinytable::tt(df)
```

&nbsp;

## Relational operators

For relational operators, 'broadcast' first finds the "highest" (or most complex) atomic type of both sides, coerces both sides to said type, and then performs the broadcasted relational operation. Recursive types are not supported.

The types, from complex to simple, are: `character`, `complex`, `numeric` (also known as `double`), `integer`, `logical`, and `raw`.


&nbsp;

# Derived operators

There is no overload for `xor()`, as `xor()` is defined using the existing base relational and logical (or bit-wise, for type of `raw`) operators.  

There is also no overload for the %d==%, %d!=%, etc. operators, as they too are defined using the existing base relational and logical operators.

&nbsp;

# Base operators vs Overloaded operators

Although the operators overloaded by 'broadcast' attempt to mimic the base `r fa("r-project")` operators accurately, the are a few differences.

Most notably, the broadcasted operators do not preserve attributes.  
Broadcasting often results in an object with more dimensions, larger dimensions, and/or larger length than the original objects.  
Therefore, the `names`, `dimnames`, and `dim` attributes often no longer fit the new object.  
Moreover, class attributes are related to dimensions or length.  
For example, the `matrix` class presumes the object to have 2 dimensions, and the various classes from the 'bit' package use length-related attributes for their functionality.  
So even class attributes cannot be guaranteed to hold for the resulting objects.  
Only some class attributes, like the 'broadcaster' class (and related) attributes, will be preserved, if present.

&nbsp;

# Overloaded operators vs `bc.*` functions

Overloading the operators is primarily useful for encuring correct mathematical precedence, and to reduce the amount of typing.  
However, the many `bc.*` functions provide much greater control than the simple operators can provide.  

For example:  
The `&` and `|` operators provide logical AND and OR, respectively for all atomic types except type `raw`; for type `raw`, the `&` and `|` operators provide BIT-WISE AND and OR, respectively.  
But what if you want to use logical AND/OR for `raw` arrays, or bit-wise AND/OR for `integer` arrays?  
When using the overloaded operators, using logical AND/OR for type `raw` necessitates converting the vector/array to type `logical`, thus making a copy.  
With the `bc.*()` functions, this is not necessary.  
Want to use logical AND/OR? Use [bc.b()](/man/bc.b.qmd), it supports several types including `raw`.  
Want to use bit-wise AND/OR? Use [bc.bit()](/man/bc.bit.qmd), it supports not only `raw`, but `integer` as well.


&nbsp;




