---
title: "Casting Explained"
format:
  html:
    number-sections: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo=FALSE, eval=TRUE}
library(broadcast) |> suppressMessages()
library(fontawesome) |> suppressWarnings()
```

# Introduction

Hierarchical data is surprisingly common, and are commonly represented in `r fa("r-project")` by nested lists.

Broadcasted operations can be performed over dimensions, but not through nesting or hierarchies.  
Therefore, it is useful to be able to cast nested lists into dimensional lists.  
The 'broadcast' package provides the [cast_hier2dim()](/man/cast_hier2dim.qmd) function, to cast nested lists into dimensional lists, and the [cast_dim2hier()](/man/cast_dim2hier.qmd) to cast dimensional lists back to nested lists.

Casting between nested and dimensional lists is not only useful for broadcasting, however.  
Casting nested lists to dimensional lists has its own merits, as dimensional lists have some advantages over nested lists beside the broadcasting, such as the following:

 - Performing sub-set operations on **multiple** recursive subsets (using the `[[` and `[[<-` operators) requires a (potentially slow) loop, whereas multi-dimensional subsets (using operator forms like `[..., ...]` and `[..., ...]<-`) are vectorized and generally much faster.
 - Re-organizing recursive arrays is generally much easier, faster, and more straight-forward than re-organizing nested lists.
 
 
This Vignette explains the functions 'broadcast' provides to cast between nested and dimensional lists.

&nbsp;


# Cast Evenly Nested List to Dimensional List

## Introduction

The "ideal" use-case for [cast_hier2dim()](/man/cast_hier2dim.qmd) is when you have an evenly nested list  - like the following list:


```{r}
x <- list(
  group1 = list(
    class1 = list(
      height = rnorm(10, 170),
      weight = rnorm(10, 80),
      sex = sample(c("M", "F", NA), 10, TRUE)
    ),
    class2 = list(
      height = rnorm(10, 170),
      weight = rnorm(10, 80),
      sex = sample(c("M", "F", NA), 10, TRUE)
    )
  ),
  group2 = list(
    class1 = list(
      height = rnorm(10, 170),
      weight = rnorm(10, 80),
      sex = sample(c("M", "F", NA), 10, TRUE)
    ),
    class2 = list(
      height = rnorm(10, 170),
      weight = rnorm(10, 80),
      sex = sample(c("M", "F", NA), 10, TRUE)
    )
  )
)

```

The above list is evenly nested, because all "group#" elements consists of the same number of "class#" elements each, and all "class#" elements consist of the same number of variables (height, weight, and sex) each.

Suppose you have 2 lists like the one above, and want to do a broadcasted operation on them. Like using some measurement to determine the difference in the elements of both lists.  
Broadcasting, as stated earlier, only works over dimensions, not through nesting.  
A list like `x` thus needs to become a dimensional list, rather than a nested list.  
And that's what the [cast_hier2dim()](/man/cast_hier2dim.qmd) function does: it re-casts a nested list to a dimensional list.

Let's try this out on list `x`.

&nbsp;

## [hier2dim()](/man/cast_hier2dim.qmd)


Before actually casting `x` into a dimensional list, one may want to know what the dimensions will become when casted as a dimensional list;  
The [hier2dim()](/man/cast_hier2dim.qmd) functions does that:

```{r}
hier2dim(x)
```

It returns `c(3, 2, 2)`.  
Notice all elements are named "no padding".  
This means that, when `x` is casted as dimensional list, no elements will be added (no "padding").  
No padding is needed because `x` is an evenly nested list.


[cast_hier2dim()](/man/cast_hier2dim.qmd) has an argument called `in2out`, which is set to `TRUE` by default.  
This means [cast_hier2dim()](/man/cast_hier2dim.qmd) will by default cast the inner most elements to the rows, and elements of the depth above that to the columns, and so on until the surface-level elements become the last dimensions.  
`x` has 3 elements at its deepest level for all iterations: height, weight, and sex; so those will become the rows.  
`x` has 2 elements at the second-deepest level for all iterations of the list: class1, class2; so those will become the columns.  
Finally `x` itself has 2 elements: group1 and group2, so those will become the last dimension (the layers).  


If the element ordering described above is not desired, one can set `in2out` to `FALSE`.  
If `in2out = FALSE`, the surface elements (in this case the "group#" elements) become the rows, the elements at one depth deeper ("class#") become the columns, and the elements yet another level deeper (height, weight, sex) become the layers (i.e. third dimension), and so on:

```{r}
hier2dim(x, in2out = FALSE)
```


## Cast in to out

Let's now cast `x` as a dimensional list:

```{r}
hier2dim(x) # check once again the dimensions
x2 <- cast_hier2dim(x) # actually cast nested list into dimensional list
print(x2)
```

`x2` is the casted list.
Since `in2out = TRUE`, element `x[[i]][[j]][[k]]` corresponds to element `x2[k, j, i]` (for all `i`, `j`, and `k`).


[cast_hier2dim()](/man/cast_hier2dim.qmd) will **not** preserve names, as there is no guarantee that the names of all elements over iterations within the same depth will have the same names.  
But in this case, it is known the names as consistent, so `x2` can very easily be given dimensional names, like so:

```{r}
# remember: in2out = TRUE
# so we go from the deepest names to the surface names
dimnames(x2) <- list(
  names(x[[1]][[1]]),
  names(x[[1]]),
  names(x)
)
print(x2)
```

There, the names are now correct.

As shown above, `r fa("r-project")` will display a dimensional list more compactly than a nested list.  
Depending on the situation this may be either be desirable or undesirable.  

One can print `x2` less compactly without much effort by flattening it, using the [cast_dim2flat()](/man/cast_dim2flat.qmd) function.  
We only need to see a portion of the list in detail, so let's look at class1 from group 1 in the flattened form:

```{r}
cast_dim2flat(x2[, 1, "group1", drop = FALSE])
```

&nbsp;

## Cast from out to in

Let's cast `x` again, but this time set the argument `in2out = FALSE`.


```{r}
hier2dim(x, in2out = FALSE) # check once again the dimensions
x2 <- cast_hier2dim(x, in2out = FALSE) # actually cast nested list into dimensional list
print(x2)
```

`x2` is the casted list.
Since `in2out = FALSE`, element `x[[i]][[j]][[k]]` corresponds to element `x2[i, j, k]` (for all `i`, `j`, and `k`).


Once again the `dimnames` need to be set manually:

```{r}
# this time, in2out = FALSE
# so we go from the surface names to the deepest names
dimnames(x2) <- list(
  names(x),
  names(x[[1]]),
  names(x[[1]][[1]])
)
print(x2)
```

There, the names are now correct.

One can print `x2` less compactly without much effort by flattening it, again using the [cast_dim2flat()](/man/cast_dim2flat.qmd) function.  
We only need to see a portion of the list in detail, so let's look at class1 from group 1 in the flattened form:

```{r}
cast_dim2flat(x2["group1", 1, , drop = FALSE])
```


&nbsp;


# Cast Unevenly Nested List to Dimensional List

In the previous section, the ideal scenario of an evenly nested list was used.   
But the [cast_hier2dim()](/man/cast_hier2dim.qmd) and [hier2dim()](/man/cast_hier2dim.qmd) functions also work on nested lists with uneven nesting.

Take for example the following list:

```{r}
x <- list(
  group1 = list(
    class1 = list(
      height = rnorm(10, 170),
      weight = rnorm(10, 80),
      sex = sample(c("M", "F", NA), 10, TRUE)
    ),
    class2 = list(
      height = rnorm(10, 170),
      weight = rnorm(10, 80),
      sex = sample(c("M", "F", NA), 10, TRUE)
    )
  ),
  group2 = list(
    class1 = list(
      height = rnorm(10, 170),
      weight = rnorm(10, 80),
      sex = sample(c("M", "F", NA), 10, TRUE)
    ),
    class2 = list(
      height = rnorm(10, 170),
      sex = sample(c("M", "F", NA), 10, TRUE)
    ),
    class3 = list(
      height = rnorm(10, 170),
      weight = rnorm(10, 80),
      sex = sample(c("M", "F", NA), 10, TRUE)
    )
  )
)

```


The above list is UNEVENLY nested:

 - not all groups (surface level, or depth = 1) have the same number of classes.
 - not all classes (depth = 2) have the same number of variables: x$group2$class2 is missing the variable "weight".

Checking what dimensions the casted list will have gives the following:

```{r}
hier2dim(x)
```

The [hier2dim()](/man/cast_hier2dim.qmd) function indicates the first 2 dimensions (rows and columns) will require some padding.

Let's cast `x`:

```{r}
x2 <- cast_hier2dim(x)
print(x2)
```

`x2` has more `NULL` values than `x`.  
The `NULL` values are added for padding, as all slices of the same dimension need to have the same number of elements.  
For example, all rows need to have the same number of columns.

[cast_dim2flat()](/man/cast_dim2flat.qmd) also works on `x2`; the additional `NULL` values used for padding do remain:

```{r}
cast_dim2flat(x2[1:2, , , drop = FALSE])
```

&nbsp;

# Cast Dimensional List to Hierarchical list

'broadcast' provides the [cast_dim2hier()](/man/cast_dim2hier.qmd):  
[cast_dim2hier()](/man/cast_dim2hier.qmd) takes a dimensional list (i.e. an array of type `list`), and casts it to a nested list.

Consider the following recursive array as an example:

```{r}

x <- array(c(as.list(1:11), ~hello, as.list(month.abb)), c(4:2))
print(x)

```


Casting it to a nested list gives the following:

```{r}

x2 <- cast_dim2hier(x)
print(x2)
           
```


Like [cast_hier2dim()](/man/cast_hier2dim.qmd), `castdim2hier()` has the `in2out` argument, which (again) defaults to `TRUE`.  
As show above, the array `x`, with dimensions `c(4, 3, 2)`, became the nested list `x2`, where `length(x2)` is `2`, `length(x2[[i]])` is `3` for every `i`, and `length(x2[[i]][[j]])` is `4` for every `i` & `j`.  
Setting `in2out = FALSE` will cast in the opposite manner:

```{r}
x2 <- cast_dim2hier(x, in2out = FALSE)
print(x2)
```

Now `length(x2)` is `4`, and the length of `length(x2[[i]][[j]])` is `2` for every `i` & `j`. 




&nbsp;

# Additional Details

## Why nested broadcasting does not work

You can perform broadcasted operations along dimensions of a dimensional list (for example using [bc.list()](/man/bc.list.qmd), [bcapply()](/man/bcapply.qmd)), but you can't really perform broadcasted operations along the depths or recursive nests of a nested list.

Broadcasting along recursive nests of a nested lists, has several issues.  
One glaring issue is that broadcasting requires matching lengths (i.e. row 1 in an array has as many columns as row 2), but matching lengths cannot be guaranteed on a nested list:  
A nested list may have, for example, 5 elements in one nested subset of a certain depth, and 4 elements on another nested subset of the same depth.  
So broadcasting can't work recursively like that.


&nbsp;

## Nested VS Dimensional Sub-set operations

Let's start with 2 lists, `x` and `y`, with essentially the same content, but `x` is a nested list, and `y` is a dimensional list:

```{r}
x <- list(
  A = list(
    A = "AA",
    B = "AB"
  ),
  B = list(
    A = "BA",
    B = "BB"
  )
)

y <- list(
  "AA", "AB", "BA", "BB"
)
dim(y) <- c(2,2)


print(x)
print(y)
```

Selecting multiple nested subsets requires a slow loop, but selecting multiple dimensional elements is vectorized and much faster.

For example, to get elements "AB" and "BB", you need to do the following:

```{r}
# for nested list x:
for(i in 1:2) {
  x[[i]][[2]] |> print()
}

# for dimensional list y:
y[2,] # much faster

```

