---
title: "List Casting Overview"
format:
  html:
    number-sections: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo=FALSE, eval=TRUE}
library(broadcast) |> suppressMessages()
library(fontawesome) |> suppressWarnings()
```

# Introduction

Hierarchical data is surprisingly common, and are commonly represented in `r fa("r-project")` by nested lists.

Broadcasted operations can be performed over dimensions, but not through nesting or hierarchies.  
Therefore, it is useful to be able to cast nested lists into dimensional lists.  
The 'broadcast' package provides the [cast_hier2dim()](/man/cast_hier2dim.qmd) function, to cast nested lists into dimensional lists, and the [cast_dim2hier()](/man/cast_dim2hier.qmd) to cast dimensional lists back to nested lists.

Casting between nested and dimensional lists is not only useful for broadcasting, however.  
Casting nested lists to dimensional lists has its own merits, as dimensional lists have some advantages over nested lists beside the broadcasting, such as the following:

 - Performing sub-set operations on **multiple** recursive subsets (using the `[[` and `[[<-` operators) requires a (potentially slow) loop, whereas multi-dimensional subsets (using operator forms like `[..., ...]` and `[..., ...]<-`) are vectorized and generally much faster.
 - Re-organizing recursive arrays is generally much easier, faster, and more straight-forward than re-organizing nested lists.
 
 
This Vignette gives an overview of the functions 'broadcast' provides to cast between nested and dimensional lists.

&nbsp;


# Cast Hierarchical List to Dimensional List

## Introduction

The [cast_hier2dim()](/man/cast_hier2dim.qmd) function casts a nested list into a dimensional list.  
This section gently introduces the properties of this function through a series of examples, where each subsequent example builds on the previous one.  
Familiarity with nested lists and dimensional lists (i.e. arrays of type `list`) is essential to follow these examples.

&nbsp;

## Example 1: Basics

For a first example, consider the following list:

```{r}
x <- list(
  group1 = list(
    class1 = list(
      height = rnorm(10, 170),
      weight = rnorm(10, 80),
      sex = sample(c("M", "F", NA), 10, TRUE)
    ),
    class2 = list(
      height = rnorm(10, 170),
      weight = rnorm(10, 80),
      sex = sample(c("M", "F", NA), 10, TRUE)
    )
  ),
  group2 = list(
    class1 = list(
      height = rnorm(10, 170),
      weight = rnorm(10, 80),
      sex = sample(c("M", "F", NA), 10, TRUE)
    ),
    class2 = list(
      height = rnorm(10, 170),
      weight = rnorm(10, 80),
      sex = sample(c("M", "F", NA), 10, TRUE)
    )
  )
)

```


Before actually casting `x` into a dimensional list, one may want to know what the dimensions will become when casted as a dimensional list;  
The [hier2dim()](/man/cast_hier2dim.qmd) function shows you that:

```{r}
hier2dim(x)
```

It returns the dimensions `c(3, 2, 2)`.  

Let's now cast `x` as a dimensional list:

```{r}
x2 <- cast_hier2dim(x) # actually cast nested list into dimensional list
print(x2)
```

Using the default arguments, element `x[[i]][[j]][[k]]` corresponds to element `x2[k, j, i]` (for all `i`, `j`, and `k`).  
This can be changed, as will be shown in a later example.


As shown in the results above, [cast_hier2dim()](/man/cast_hier2dim.qmd) will obviously not preserve names.

In this case, it is trivially easy to set the `dimnames` of `x2`:

```{r}
# remember: element `x[[i]][[j]][[k]]` corresponds to element `x2[k, j, i]`
# so we go from the deepest names to the surface names
dimnames(x2) <- list(
  names(x[[1]][[1]]),
  names(x[[1]]),
  names(x)
)
print(x2)
```

There, the names are now correct.

As shown above, `r fa("r-project")` will display a dimensional list more compactly than a nested list.  
Depending on the situation this may be either be desirable or undesirable.  

One can print `x2` less compactly without much effort by flattening it, using the [cast_dim2flat()](/man/cast_dim2flat.qmd) function.  
We only need to see a portion of the list in detail, so let's look at class1 from group 1 in the flattened form:

```{r}
cast_dim2flat(x2[, 1, "group1", drop = FALSE])
```


&nbsp;

## Example 2: Cast from outside to inside


In Example 1, the default arguments were used for [cast_hier2dim()](/man/cast_hier2dim.qmd).  
One of these arguments is `in2out`, which defaults to `TRUE`.

Consider a nested list `x` with a depth of 3, and a dimensional list `X2` with 3 dimensions, where the relationship between `x` and `x2` can be expressed as `x2 <- cast_hier2dim(x, ...)`.  
Given this, the following can be stated about `in2out`:

 - If `in2out = TRUE`, which is the default and used in Example 1, element `x[[i]][[j]][[k]]` corresponds to element `x2[k, j, i]` (for all `i`, `j`, and `k`).  
 - If `in2out = FALSE`, element `x[[i]][[j]][[k]]` corresponds to element `x2[i, j, k]` (for all `i`, `j`, and `k`).  

The default of `in2out = TRUE` was chosen, because elements in subsequent rows are close to each other, while elements in subsequent layers (third dimension) are generally not close to each other, and the default of `in2out = TRUE` attempts to retain that behaviour.

For this example, the same list will be used as in Example 1:

```{r}
x <- list(
  group1 = list(
    class1 = list(
      height = rnorm(10, 170),
      weight = rnorm(10, 80),
      sex = sample(c("M", "F", NA), 10, TRUE)
    ),
    class2 = list(
      height = rnorm(10, 170),
      weight = rnorm(10, 80),
      sex = sample(c("M", "F", NA), 10, TRUE)
    )
  ),
  group2 = list(
    class1 = list(
      height = rnorm(10, 170),
      weight = rnorm(10, 80),
      sex = sample(c("M", "F", NA), 10, TRUE)
    ),
    class2 = list(
      height = rnorm(10, 170),
      weight = rnorm(10, 80),
      sex = sample(c("M", "F", NA), 10, TRUE)
    )
  )
)

```


Let's once again cast this list to a dimensional list, but this time use `in2out = FALSE`:


```{r}
hier2dim(x, in2out = FALSE) # check once again the dimensions
x2 <- cast_hier2dim(x, in2out = FALSE) # actually cast nested list into dimensional list
print(x2)
```

`x2` is the casted list.
Since `in2out = FALSE`, element `x[[i]][[j]][[k]]` corresponds to element `x2[i, j, k]` (for all `i`, `j`, and `k`).


Once again the `dimnames` need to be set manually:

```{r}
# this time, in2out = FALSE
# so we go from the surface names to the deepest names
dimnames(x2) <- list(
  names(x),
  names(x[[1]]),
  names(x[[1]][[1]])
)
print(x2)
```

There, the names are now correct.

One can print `x2` less compactly without much effort by flattening it, again using the [cast_dim2flat()](/man/cast_dim2flat.qmd) function.  
We only need to see a portion of the list in detail, so let's look at class1 from group 1 in the flattened form:

```{r}
cast_dim2flat(x2["group1", 1, , drop = FALSE])
```

&nbsp;


## Example 3: Padding

For Example 3, we take the same list as before, but remove `x$group1$class2`:

```{r}
x <- list(
  group1 = list(
    class1 = list(
      height = rnorm(10, 170),
      weight = rnorm(10, 80),
      sex = sample(c("M", "F", NA), 10, TRUE)
    )
  ),
  group2 = list(
    class1 = list(
      height = rnorm(10, 170),
      weight = rnorm(10, 80),
      sex = sample(c("M", "F", NA), 10, TRUE)
    ),
    class2 = list(
      height = rnorm(10, 170),
      weight = rnorm(10, 80),
      sex = sample(c("M", "F", NA), 10, TRUE)
    )
  )
)


```

Let's first check what dimensions it will get when casted using [hier2dim()](/man/cast_hier2dim.qmd):

```{r}
hier2dim(x)
```

The dimensions are the same as in Example 1: `c(3, 2, 2)`.  
But notice the **names** of the output are different: the second element has the name "padding"; this indicates that some columns won't have enough elements to completely fill the column, and so additional elements will be added as padding.

So let's cast this list as dimensional:

```{r}
x2 <- cast_hier2dim(x)
print(x2)
```

Subset `x2[, 2, 1]` is filled with `NULL`; this is the place where `x$group1$class2` was in Example 1, but since it's not there, we need to fill **something**.

To make it make obvious, let's give the array proper `dimnames`:

```{r}
dimnames(x2) <- list(
  c("height", "weight", "sex"),
  c("class1", "class2"),
  c("group1", "group2")
)
print(x2)
```

Again, element "class2" is missing from element "group1", but not from "group2", and so padded with `NULL` when the list is casted as dimensional.

Sometimes, a different value than `NULL` is desired for padding.  
So let's replace the padding value with something **really** obvious, using the `padding` argument:

```{r}
x2 <- cast_hier2dim(x, padding = list(~ "this is padding!"))
dimnames(x2) <- list(
  c("height", "weight", "sex"),
  c("class1", "class2"),
  c("group1", "group2")
)
print(x2)
```

Once again, one can print or present `x2` less compactly by flattening it:

```{r}
cast_dim2flat(x2)
```

&nbsp;

## Example 4: Comparing `in2out` with padding

In this example, the same nested list as from the previous example is used, to demonstrate the difference between `in2out = TRUE` (which is the default), and `in2out = FALSE`.

On the left side the original nested list is shown.  
On the right side, the list casted as dimensional list (using the default `in2out = TRUE`), with the proper names assigned, is shown:

:::{layout-ncol="2"}

```{r}
x <- list(
  group1 = list(
    class1 = list(
      height = rnorm(10, 170),
      weight = rnorm(10, 80),
      sex = sample(c("M", "F", NA), 10, TRUE)
    )
  ),
  group2 = list(
    class1 = list(
      height = rnorm(10, 170),
      weight = rnorm(10, 80),
      sex = sample(c("M", "F", NA), 10, TRUE)
    ),
    class2 = list(
      height = rnorm(10, 170),
      weight = rnorm(10, 80),
      sex = sample(c("M", "F", NA), 10, TRUE)
    )
  )
)


```

```{r}
x2 <- cast_hier2dim(x)
dimnames(x2) <- list(
  c("height", "weight", "sex"),
  c("class1", "class2"),
  c("group1", "group2")
)
print(x2)
```

:::


&nbsp;

Now we do the same, but with `in2out = FALSE`.  
On the left side the original nested list is shown.  
On the right side, the list casted as dimensional list - using `in2out = FALSE`, with the proper names assigned, is shown:

:::{layout-ncol="2"}

```{r}
x <- list(
  group1 = list(
    class1 = list(
      height = rnorm(10, 170),
      weight = rnorm(10, 80),
      sex = sample(c("M", "F", NA), 10, TRUE)
    )
  ),
  group2 = list(
    class1 = list(
      height = rnorm(10, 170),
      weight = rnorm(10, 80),
      sex = sample(c("M", "F", NA), 10, TRUE)
    ),
    class2 = list(
      height = rnorm(10, 170),
      weight = rnorm(10, 80),
      sex = sample(c("M", "F", NA), 10, TRUE)
    )
  )
)


```

```{r}
x2 <- cast_hier2dim(x, in2out = FALSE)
dimnames(x2) <- list(
  c("group1", "group2"),
  c("class1", "class2"),
  c("height", "weight", "sex")
)
print(x2)
```

:::

&nbsp;

# Cast Dimensional List to Hierarchical list

'broadcast' provides the [cast_dim2hier()](/man/cast_dim2hier.qmd):  
[cast_dim2hier()](/man/cast_dim2hier.qmd) takes a dimensional list (i.e. an array of type `list`), and casts it to a nested list.

Consider the following recursive array as an example:

```{r}

x <- array(c(as.list(1:11), ~hello, as.list(month.abb)), c(4:2))
dimnames(x) <- list(
  letters[1:4],
  LETTERS[1:3],
  c("group1", "group2")
)
print(x)

```


Like [cast_hier2dim()](/man/cast_hier2dim.qmd), [cast_dim2hier()](/man/cast_dim2hier.qmd) also has the `in2out` argument, which (again) defaults to `TRUE`.  
Let's cast the above dimensional list to a nested list, and compare the results when using `in2out = TRUE` (on the left) versus `in2out = FALSE` (on the right):


:::{layout-ncol="2"}

```{r}

x2 <- cast_dim2hier(
  x, distr.names = TRUE
)
lobstr::tree(x2)
           
```


```{r}

x2 <- cast_dim2hier(
  x, in2out = FALSE, distr.names = TRUE
)
lobstr::tree(x2)
           
```

:::


The added `distr.names = TRUE` argument will distribute the `dimnames` in a logical way over the nested elements.



&nbsp;

&nbsp;

