---
title: "Casting Explained"
format:
  html:
    number-sections: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo=FALSE, eval=TRUE}
library(broadcast) |> suppressMessages()
library(fontawesome) |> suppressWarnings()
```


# Cast Hierarchical List to Dimensional List

## Introduction

The 'broadcast' package provides the [cast_hier2dim()](/man/cast_hier2dim.qmd) function, to cast nested lists into dimensional lists.  
This is useful as dimensional list have 2 (relevant for now) advantages over nested lists without dimensions:

 - Broadcasted operations can be performed over dimensions, but not through nesting or hierarchies;
 - Performing sub-set operations on **multiple** recursive subsets (using the `[[` and `[[<-` operators) requires a loop, whereas multi-dimensional subsets (using operator forms like `[..., ...]` and `[..., ...]<-`) are vectorized

&nbsp;

## Example of ideal scenario

The "ideal" use-case for [cast_hier2dim()](/man/cast_hier2dim.qmd) is when you have a list where all elements within the same depth have equal number of elements - like the following lists:

```{r}
x <- list(
  group1 = list(
    id1 = list(
      height = rnorm(10, 170),
      weight = rnorm(10, 80),
      sex = sample(c("M", "F", NA), 10, TRUE)
    ),
    id2 = list(
      height = rnorm(10, 170),
      weight = rnorm(10, 80),
      sex = sample(c("M", "F", NA), 10, TRUE)
    ),
    id3 = list(
      height = rnorm(10, 170),
      weight = rnorm(10, 80),
      sex = sample(c("M", "F", NA), 10, TRUE)
    )
  ),
  group2 = list(
    id1 = list(
      height = rnorm(10, 170),
      weight = rnorm(10, 80),
      sex = sample(c("M", "F", NA), 10, TRUE)
    ),
    id2 = list(
      height = rnorm(10, 170),
      weight = rnorm(10, 80),
      sex = sample(c("M", "F", NA), 10, TRUE)
    ),
    id3 = list(
      height = rnorm(10, 170),
      weight = rnorm(10, 80),
      sex = sample(c("M", "F", NA), 10, TRUE)
    )
  )
)

y <- list(
  group1 = list(
    id1 = list(
      height = rnorm(10, 170),
      weight = rnorm(10, 80),
      sex = sample(c("M", "F", NA), 10, TRUE)
    ),
    id2 = list(
      height = rnorm(10, 170),
      weight = rnorm(10, 80),
      sex = sample(c("M", "F", NA), 10, TRUE)
    ),
    id3 = list(
      height = rnorm(10, 170),
      weight = rnorm(10, 80),
      sex = sample(c("M", "F", NA), 10, TRUE)
    )
  ),
  group2 = list(
    id1 = list(
      height = rnorm(10, 170),
      weight = rnorm(10, 80),
      sex = sample(c("M", "F", NA), 10, TRUE)
    ),
    id2 = list(
      height = rnorm(10, 170),
      weight = rnorm(10, 80),
      sex = sample(c("M", "F", NA), 10, TRUE)
    ),
    id3 = list(
      height = rnorm(10, 170),
      weight = rnorm(10, 80),
      sex = sample(c("M", "F", NA), 10, TRUE)
    )
  )
)
```

Suppose you have 2 lists like the above, and want to do a broadcasted operation on them. Like using some measurement to determine the difference in the elements of both lists.  
Broadcasting, as stated earlier, only works over dimensions, not through nesting.  
Lists `x` and `y` thus need to become dimensional lists, rather than nested lists.  
And that's what the [cast_hier2dim()](/man/cast_hier2dim.qmd) function does: it re-casts a nested list to a dimensional list.

Let's try this out on list `x`.

Before actually casting `x` into a dimensional list, one may want to know what the dimensions will become when casted as a dimensional list;  
The [hier2dim()](/man/cast_hier2dim.qmd) functions does that:

```{r}
hier2dim(x)
```

It returns `c(3, 3, 2)`.  
`cast_hier2dim` has an argument called `in2out`, which is set to `TRUE` by default.  
This means [cast_hier2dim()](/man/cast_hier2dim.qmd) will by default cast the inner most elements to the rows, and elements of the depth above that to the columns, and so on until the surface-level elements become the last dimensions.  
`x` has 3 elements at its deepest level for all iterations: height, weight, and sex; so those will become the rows.  
`x` also has 3 elements at the second-deepest level for all iterations of the list: id1, id2, id3; so those will become the columns.  
Finally `x` itself has 2 elements: group1 and group2, so those will become the last dimension (the layers).  


If the element ordering described above is not desired, one can set `in2out` to `FALSE`.  
If `in2out = FALSE`, the surface elements become the rows, the elements at one depth deeper the columns, and the elements yet another level deeper the layers (i.e. third dimension), and so on:

```{r}
hier2dim(x, in2out = FALSE)
```


For this Vignette, `in2out` is left at its default value of `TRUE`.  
Let's now cast `x` as a dimensional list:

```{r}
hier2dim(x) # check once again the dimensions
x2 <- cast_hier2dim(x) # actually cast nested list into dimensional list
print(x2)
```

`x2` is the casted list.
Since `in2out = TRUE`, element `x[[i]][[j]][[k]]` corresponds to element `x2[k, j, i]` (for all `i`, `j`, and `k`).
If `in2out` were set to `FALSE`, element `x[[i]][[j]][[k]]` would correspond to element `x2[i, j, k]` (for all `i`, `j`, and `k`).


[cast_hier2dim()](/man/cast_hier2dim.qmd) will **not** preserve names, as there is no guarantee that the names of all elements over iterations within the same depth will have the same names.  
But in this case, it is known the names as consistent, so `x2` can very easily be given dimensional names, like so:

```{r}
# remember: in2out = TRUE
# so we go from the deepest names to the surface names
dimnames(x2) <- list(
  names(x[[1]][[1]]),
  names(x[[1]]),
  names(x)
)
print(x2)
```

There, the names are now correct.

As shown above, 'R' will display a dimensional list more compactly than a nested list.  
Depending on the situation this may be either be desirable or undesirable.  

One can print `x2` less compactly without much effort by flattening it.  
We only need to see a portion of the list in detail, so let's look at id1 and id2 from group 1 in the flattened form:

```{r}
cast_dim2flat(x2[, 1:2, "group1", drop = FALSE])
```

[cast_dim2flat()](/man/cast_dim2flat.qmd) function.

&nbsp;

## Less ideal scenario

Up until now an ideal scenario was used.  
I.e. within the same depth, all iterations had the same number of elements.  
But the [cast_hier2dim()](/man/cast_hier2dim.qmd) and [hier2dim()](/man/cast_hier2dim.qmd) functions also work on nested lists with uneven nesting.

Take for example the following list:

```{r}
x <- list(
  A = list(
    A = list(A = "AAA", B = "AAB", list(NULL)),
    A = list(A  = "AA2A", B = "AA2B", list(NA)),
    B = list(A = "ABA", B = "ABB", list(), list(NULL))
  ),
  B = list(
    A = list(A = "BAA", B = "BAB", NULL),
    B = list(A = "BBA", B = "BBB", list(NA)),
    B = list(A = "BB2A", B = "BB2B", list())
  )
)
```


Let's see what dimensions one would get if this  nested list were to be made dimensional:

```{r}
hier2dim(x)
```

We get `c(4, 3, 2)`.  
`x` has 2 elements, so that explains the `2` at the end.  
Every element `x[[i]]` has 3 elements, so that explains the `3` in the middle.
At its deepest level, `x` usually has 3 elements, but at one spot - namely `x$A$B` - it has 4 elements. So therefor the first dimension becomes `4`. 


Let's cast `x`:

```{r}
x2 <- cast_hier2dim(x)
print(x2)
```

`x2` has more `NULL` values than `x`:  
Remember that the dimensions take the maximum sizes, which means some rows/columns/etc. need to be padded; `NULL` is used for the padding.

As the original list `x` is unevenly nested, there is little point in giving the resulting list dimensional names.

[cast_dim2flat()](/man/cast_dim2flat.qmd) also works on `x2`; the additional `NULL` values used for padding do remain:

```{r}
cast_dim2flat(x2[, 1:2, , drop = FALSE])
```


&nbsp;

## Additional Details

### Why nested broadcasting does not work

You can perform broadcasted operations along dimensions of a dimensional list (for example using [bc.list()](/man/bc.list.qmd), [bcapply()](/man/bcapply.qmd)), but you can't really perform broadcasted operations along the depths or recursive nests of a nested list.

Broadcasting along recursive nests of a nested lists, has several issues.  
One glaring issue is that broadcasting requires matching lengths (i.e. row 1 in an array has as many columns as row 2), but matching lengths cannot be guaranteed on a nested list:  
A nested list may have, for example, 5 elements in one nested subset of a certain depth, and 4 elements on another nested subset of the same depth.  
So broadcasting can't work recursively like that.

One easy solution is to take a nested list, which could conceivably be represented as a dimensional list, and then actually cast the nested list as a dimensional list.  
And this is what the [cast_hier2dim()](/man/cast_hier2dim.qmd) function does.  

'broadcast' provides the [cast_hier2dim()](/man/cast_hier2dim.qmd) function, to cast nested lists into dimensional lists.

&nbsp;

### Nested VS Dimensional Sub-set operations

Let's start with 2 lists, `x` and `y`, with essentially the same content, but `x` is a nested list, and `y` is a dimensional list:

```{r}
x <- list(
  A = list(
    A = "AA",
    B = "AB"
  ),
  B = list(
    A = "BA",
    B = "BB"
  )
)

y <- list(
  "AA", "AB", "BA", "BB"
)
dim(y) <- c(2,2)


print(x)
print(y)
```

Selecting multiple nested subsets requires a slow loop, but selecting multiple dimensional elements is vectorized and much faster.

For example, to get elements "AB" and "BB", you need to do the following:

```{r}
# for x:
for(i in 1:2) {
  x[[i]][[2]] |> print()
}

# for y:
y[2,] # much faster

```


&nbsp;

# Cast Dimensional List to Hierarchical list

'broadcast' provides the [cast_dim2hier()](/man/cast_dim2hier.qmd):  
[cast_dim2hier()](/man/cast_dim2hier.qmd) takes a dimensional list (i.e. an array of type `list`), and casts it to a nested list.

For example:

```{r}

x <- array(c(as.list(1:11), ~hello, as.list(month.abb)), c(4:2))
print(x)

x2 <- cast_dim2hier(x)
print(x2)
           
```


Like [cast_hier2dim()](/man/cast_hier2dim.qmd), `castdim2hier()` has the `in2out` argument, which (again) defaults to `TRUE`.  
As show above, the array `x`, with dimensions `c(4, 3, 2)`, became the nested list `x2`, where `length(x2)` is `2`, `length(x2[[i]])` is `3` for every `i`, and `length(x2[[i]][[j]])` is `4` for every `i` & `j`.  
Setting `in2out = FALSE` will cast in the opposite manner:

```{r}
x2 <- cast_dim2hier(x, in2out = FALSE)
print(x2)
```

Now `length(x2)` is `4`, and the length of `length(x2[[i]][[j]])` is `2` for every `i` & `j`. 


