---
title: "Casting Explained"
format:
  html:
    number-sections: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo=FALSE, eval=TRUE}
library(broadcast) |> suppressMessages()
library(fontawesome) |> suppressWarnings()
```

# Introduction

Hierarchical data is surprisingly common, and are commonly represented in `r fa("r-project")` by nested lists.

Broadcasted operations can be performed over dimensions, but not through nesting or hierarchies.  
Therefore, it is useful to be able to cast nested lists into dimensional lists.  
The 'broadcast' package provides the [cast_hier2dim()](/man/cast_hier2dim.qmd) function, to cast nested lists into dimensional lists, and the [cast_dim2hier()](/man/cast_dim2hier.qmd) to cast dimensional lists back to nested lists.

Casting between nested and dimensional lists is not only useful for broadcasting, however.  
Casting nested lists to dimensional lists has its own merits, as dimensional lists have some advantages over nested lists beside the broadcasting, such as the following:

 - Performing sub-set operations on **multiple** recursive subsets (using the `[[` and `[[<-` operators) requires a (potentially slow) loop, whereas multi-dimensional subsets (using operator forms like `[..., ...]` and `[..., ...]<-`) are vectorized and generally much faster.
 - Re-organizing recursive arrays is generally much easier, faster, and more straight-forward than re-organizing nested lists.
 
 
This Vignette explains the functions 'broadcast' provides to cast between nested and dimensional lists.

&nbsp;


# Cast Nested List to Dimensional List

## Introduction

The [cast_hier2dim()](/man/cast_hier2dim.qmd) function casts a nested list into a dimensional list.  
In this section, the properties of this function is shown through a series of examples, where the complexity of each subsequent example increases gently.

&nbsp;

## Example 1: Basics

For a first example, consider the following list:

```{r}
x <- list(
  group1 = list(
    class1 = list(
      height = rnorm(10, 170),
      weight = rnorm(10, 80),
      sex = sample(c("M", "F", NA), 10, TRUE)
    ),
    class2 = list(
      height = rnorm(10, 170),
      weight = rnorm(10, 80),
      sex = sample(c("M", "F", NA), 10, TRUE)
    )
  ),
  group2 = list(
    class1 = list(
      height = rnorm(10, 170),
      weight = rnorm(10, 80),
      sex = sample(c("M", "F", NA), 10, TRUE)
    ),
    class2 = list(
      height = rnorm(10, 170),
      weight = rnorm(10, 80),
      sex = sample(c("M", "F", NA), 10, TRUE)
    )
  )
)

```


Before actually casting `x` into a dimensional list, one may want to know what the dimensions will become when casted as a dimensional list;  
The [hier2dim()](/man/cast_hier2dim.qmd) function shows you that:

```{r}
hier2dim(x)
```

It returns the dimensions `c(3, 2, 2)`.  

Let's now cast `x` as a dimensional list:

```{r}
x2 <- cast_hier2dim(x) # actually cast nested list into dimensional list
print(x2)
```

Using the default arguments, element `x[[i]][[j]][[k]]` corresponds to element `x2[k, j, i]` (for all `i`, `j`, and `k`).  
This can be changed, as will be shown in a later example.


As shown in the results above, [cast_hier2dim()](/man/cast_hier2dim.qmd) will **not** preserve names:

In arrays, all rows share the same row names, all columns share the same column names and so on. Moreover, all rows have the same number of columns (and vice-versa) and so on. But these things are not necessarily the case for nested lists. And so  [cast_hier2dim()](/man/cast_hier2dim.qmd) won't know how to name the elements.

But in this case, it is known the names are consistent, so `x2` can very easily be given dimensional names, like so:

```{r}
# remember: element `x[[i]][[j]][[k]]` corresponds to element `x2[k, j, i]`
# so we go from the deepest names to the surface names
dimnames(x2) <- list(
  names(x[[1]][[1]]),
  names(x[[1]]),
  names(x)
)
print(x2)
```

There, the names are now correct.

As shown above, `r fa("r-project")` will display a dimensional list more compactly than a nested list.  
Depending on the situation this may be either be desirable or undesirable.  

One can print `x2` less compactly without much effort by flattening it, using the [cast_dim2flat()](/man/cast_dim2flat.qmd) function.  
We only need to see a portion of the list in detail, so let's look at class1 from group 1 in the flattened form:

```{r}
cast_dim2flat(x2[, 1, "group1", drop = FALSE])
```


&nbsp;

## Example 2: Cast from outside to inside


In Example 1, the default arguments were used for [cast_hier2dim()](/man/cast_hier2dim.qmd).  
One of these arguments is `in2out`, which defaults to `TRUE`.

Consider a nested list `x` with a depth of 3, and a dimensional list `X2` with 3 dimensions, where the relationship between `x` and `x2` can be expressed as `x2 <- cast_hier2dim(x, ...)`.  
Given this, the following can be stated about `in2out`:

 - If `in2out = TRUE`, which is the default and used in Example 1, element `x[[i]][[j]][[k]]` corresponds to element `x2[k, j, i]` (for all `i`, `j`, and `k`).  
 - If `in2out = FALSE`, element `x[[i]][[j]][[k]]` corresponds to element `x2[i, j, k]` (for all `i`, `j`, and `k`).  

The default of `in2out = TRUE` was chosen, because elements in subsequent rows are close to each other, while elements in subsequent layers (third dimension) are generally not close to each other, and [cast_hier2dim()](/man/cast_hier2dim.qmd) attempts to retain that behaviour.

For this example, the same list will be used as in Example 1:

```{r}
x <- list(
  group1 = list(
    class1 = list(
      height = rnorm(10, 170),
      weight = rnorm(10, 80),
      sex = sample(c("M", "F", NA), 10, TRUE)
    ),
    class2 = list(
      height = rnorm(10, 170),
      weight = rnorm(10, 80),
      sex = sample(c("M", "F", NA), 10, TRUE)
    )
  ),
  group2 = list(
    class1 = list(
      height = rnorm(10, 170),
      weight = rnorm(10, 80),
      sex = sample(c("M", "F", NA), 10, TRUE)
    ),
    class2 = list(
      height = rnorm(10, 170),
      weight = rnorm(10, 80),
      sex = sample(c("M", "F", NA), 10, TRUE)
    )
  )
)

```


Let's once again cast this list to a dimensional list, but this time use `in2out = FALSE`:


```{r}
hier2dim(x, in2out = FALSE) # check once again the dimensions
x2 <- cast_hier2dim(x, in2out = FALSE) # actually cast nested list into dimensional list
print(x2)
```

`x2` is the casted list.
Since `in2out = FALSE`, element `x[[i]][[j]][[k]]` corresponds to element `x2[i, j, k]` (for all `i`, `j`, and `k`).


Once again the `dimnames` need to be set manually:

```{r}
# this time, in2out = FALSE
# so we go from the surface names to the deepest names
dimnames(x2) <- list(
  names(x),
  names(x[[1]]),
  names(x[[1]][[1]])
)
print(x2)
```

There, the names are now correct.

One can print `x2` less compactly without much effort by flattening it, again using the [cast_dim2flat()](/man/cast_dim2flat.qmd) function.  
We only need to see a portion of the list in detail, so let's look at class1 from group 1 in the flattened form:

```{r}
cast_dim2flat(x2["group1", 1, , drop = FALSE])
```

&nbsp;

## Example 3: Padding

For Example 3, we take the same list as before, but remove `x$group1$class2`:

```{r}
x <- list(
  group1 = list(
    class1 = list(
      height = rnorm(10, 170),
      weight = rnorm(10, 80),
      sex = sample(c("M", "F", NA), 10, TRUE)
    )
  ),
  group2 = list(
    class1 = list(
      height = rnorm(10, 170),
      weight = rnorm(10, 80),
      sex = sample(c("M", "F", NA), 10, TRUE)
    ),
    class2 = list(
      height = rnorm(10, 170),
      weight = rnorm(10, 80),
      sex = sample(c("M", "F", NA), 10, TRUE)
    )
  )
)


```

Let's first check what dimensions it will get when casted using [hier2dim()](/man/cast_hier2dim.qmd):

```{r}
hier2dim(x)
```

The dimensions are the same as in Example 1: `c(3, 2, 2)`.  
But notice the **names** of the output are different: the second element has the name "padding", instead of "no padding".  
This indicates that some columns won't have enough elements to completely fill the column, and so additional elements will be added as padding.

Because remember: in an array, all rows must have the same number of columns and vice-versa, all columns must have the same number of layers, and so on.

So let's cast this list as dimensional:

```{r}
x2 <- cast_hier2dim(x)
print(x2)
```

Subset `x2[1, , 2]` is filled with `NULL`; this is the place where `x$group1$class2` was in Example 1, but since it's not there, we need to fill **something**.

To make it make obvious, let's give the list proper names:

```{r}
dimnames(x2) <- list(
  c("height", "weight", "sex"),
  c("class1", "class2"),
  c("group1", "group2")
)
print(x2)
```

Again, element "class2" is missing from element "group1", but not from "group2", and so padded with `NULL` when the list is casted as dimensional.

Sometimes, a different value than `NULL` is desired for padding.  
For example, when casting `x2` back to a nested list, or when casting `x2` to a flat list, I can imagine you'd want to disregard the padded values.  
Padding with `NULL` may then be inconvenient, as a list may contain "true" `NULL` values.

So let's change the padding with something REALLY obvious, using the `padding` argument:

```{r}
x2 <- cast_hier2dim(x, padding = list(~ "this is padding!"))
dimnames(x2) <- list(
  c("height", "weight", "sex"),
  c("class1", "class2"),
  c("group1", "group2")
)
print(x2)
```

Once again, one can print or summarise `x2` less compactly by flattening it:

```{r}
cast_dim2flat(x2)
```


&nbsp;

# Cast Dimensional List to Hierarchical list

'broadcast' provides the [cast_dim2hier()](/man/cast_dim2hier.qmd):  
[cast_dim2hier()](/man/cast_dim2hier.qmd) takes a dimensional list (i.e. an array of type `list`), and casts it to a nested list.

Consider the following recursive array as an example:

```{r}

x <- array(c(as.list(1:11), ~hello, as.list(month.abb)), c(4:2))
dimnames(x) <- list(
  letters[1:4],
  LETTERS[1:3],
  c("group1", "group2")
)
print(x)

```


Casting it to a nested list gives the following:

```{r}

x2 <- cast_dim2hier(x, distr.names = TRUE)
print(x2)
           
```

The added `distr.names = TRUE` argument will distribute the `dimnames` in a logical way over the nested elements.

Like [cast_hier2dim()](/man/cast_hier2dim.qmd), `castdim2hier()` has the `in2out` argument, which (again) defaults to `TRUE`.  
As show above, the array `x`, with dimensions `c(4, 3, 2)`, became the nested list `x2`, where `length(x2)` is `2`, `length(x2[[i]])` is `3` for every `i`, and `length(x2[[i]][[j]])` is `4` for every `i` & `j`.  
Setting `in2out = FALSE` will cast in the opposite manner:

```{r}
x2 <- cast_dim2hier(x, in2out = FALSE, distr.names = TRUE)
print(x2)
```

Now `length(x2)` is `4`, and the length of `length(x2[[i]][[j]])` is `2` for every `i` & `j`. 




&nbsp;

# Additional Details

## Why nested broadcasting does not work

You can perform broadcasted operations along dimensions of a dimensional list (for example using [bc.list()](/man/bc.list.qmd), [bcapply()](/man/bcapply.qmd)), but you can't really perform broadcasted operations along the depths or recursive nests of a nested list.

Broadcasting along recursive nests of a nested lists, has several issues.  
One glaring issue is that broadcasting requires matching lengths (i.e. row 1 in an array has as many columns as row 2), but matching lengths cannot be guaranteed on a nested list:  
A nested list may have, for example, 5 elements in one nested subset of a certain depth, and 4 elements on another nested subset of the same depth.  
So broadcasting can't work recursively like that.


&nbsp;

## Nested VS Dimensional Sub-set operations

Let's start with 2 lists, `x` and `y`, with essentially the same content, but `x` is a nested list, and `y` is a dimensional list:

```{r}
x <- list(
  A = list(
    A = "AA",
    B = "AB"
  ),
  B = list(
    A = "BA",
    B = "BB"
  )
)

y <- list(
  "AA", "AB", "BA", "BB"
)
dim(y) <- c(2,2)


print(x)
print(y)
```

Selecting multiple nested subsets requires a slow loop, but selecting multiple dimensional elements is vectorized and much faster.

For example, to get elements "AB" and "BB", you need to do the following:

```{r}
# for nested list x:
for(i in 1:2) {
  x[[i]][[2]] |> print()
}

# for dimensional list y:
y[2,] # much faster

```

