---
title: "Prerequisites & Installation"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(fontawesome)
```

# Prerequisites

First, a basic understanding of `r fa("r-project")` is important.  
A very basic refresher for `r fa("r-project")` can be found in the [Getting Started in R: Tinyverse Edition](https://eddelbuettel.github.io/gsir-te/Getting-Started-in-R.pdf) document.

&nbsp;

Make sure you at least have basic knowledge of the following subjects in `r fa("r-project")`:

 - Atomic types (logical, integer, double, complex, character, raw).
 - The recursive type "list".
 - dimensional structures: vectors, matrices, arrays.
 - common arithmetic operators (+, -, *, /, etc.), relational operators (==, !=, etc.), logical operators (&, |), and common bit-wise operations.
 - vector recycling

&nbsp;

# Requirements

'broadcast' requires `r fa("r-project")` version 4.2.0 or later; you can download `r fa("r-project")` from [https://cran.r-project.org/mirrors.html](https://cran.r-project.org/mirrors.html).  
'broadcast' has only one dependency: 'Rcpp' version 1.0.14 or later. To install 'Rcpp', run `install.packages("Rcpp")` in `r fa("r-project")`.  

Those using the Windows Operating System also need 'RTools'. You can download 'RTools' from [https://cran.r-project.org/bin/windows/Rtools/](https://cran.r-project.org/bin/windows/Rtools/).

&nbsp;

# Installation

## Basic Instructions

To install the current version of 'broadcast' from CRAN, run the following code in `r fa("r-project")`:

```{r echo=TRUE, eval=FALSE}
install.packages("broadcast", type = "source")
```

It is possible that you may be asked if you want to install from sources that require compilation; select "Yes"; note that this requires 'RTools'. 


To install the development version of 'broadcast' from GitHub, one may run the following code in `r fa("r-project")`:

```{r echo=TRUE, eval=FALSE}

remotes::install_github("https://github.com/tony-aw/broadcast")

```


&nbsp;

## Additional Information for Advanced Users.

The following instructions are only for advanced users.

First, 'broadcast' will be compiled as instructed by the user's own 'Makevars' / 'Makevars.win' file - if it exists. It might be possible to make the methods and functions of the 'broadcast' package *faster* by adding the `CXXFLAGS` compiler flag `-O3` to your 'Makevars' / 'Makevars.win' file, which enables more aggressive optimization during installation.  
The Makevars file should be located/created in the `path.expand("~/.R")` directory.

&nbsp;

Second, 'broadcast' is not a large package per-sé, but it does contain a large amount of compiled code ('C' and 'C++' code). Compilation of the package, especially with `-O3`, may therefore be slow on some operating systems, especially Linux systems; please be patient. On an average PC running Windows 11, installation of the 'broadcast' package shouldn't take more than just a few minutes.  
If installation is too slow, one may try to speed up the installation using multi-threading, by running the following code before installing 'broadcast':

```{r echo=TRUE, eval=FALSE}

nthreads <- 8L # specify number of threads
Sys.setenv(MAKEFLAGS = paste0("-j", nthreads))


```


&nbsp;

# Start using the package

To attach the 'broadcast' package - thus expose its functions to the namespace - please use the following:

```{r echo=TRUE, eval=FALSE}

library(broadcast)

```


&nbsp;

To open the introduction help page of the ‘broadcast’ package, please use any of the following:

```{r echo=TRUE, eval=FALSE}

?broadcast::broadcast
?broadcast::broadcast_help
?broadcast::`broadcast-package`

```


&nbsp;
&nbsp;
